Loss of PTPMT1 limits mitochondrial utilization of carbohydrates and leads to muscle atrophy and heart failure

Version 1

Abstract

Abstract

While mitochondria in different tissues have distinct preferences for energy sources, they are flexible in utilizing competing substrates for metabolism according to physiological and nutritional circumstances. However, the regulatory mechanisms and significance of metabolic flexibility are not completely understood. Here we report that the deletion of PTPMT1, a mitochondria-based phosphatase, critically alters mitochondrial fuel selection – the utilization of pyruvate, a key mitochondrial substrate derived from glucose (the major simple carbohydrate), is inhibited, whereas the fatty acid utilization is enhanced. PTPMT1 knockout does not impact the development of the skeletal muscle or heart. However, the metabolic inflexibility ultimately leads to muscular atrophy, heart failure, and sudden death. Mechanistic analyses reveal that the prolonged substrate shift from carbohydrates to lipids causes oxidative stress and mitochondrial destruction, which in turn results in marked accumulation of lipids and profound damage in the knockout muscle cells and cardiomyocytes. Interestingly, PTPMT1 deletion from the liver or adipose tissue does not generate any local or systemic defects. These findings suggest that PTPMT1 plays an important role in maintaining mitochondrial flexibility and that their balanced utilization of carbohydrates and lipids is essential for both the skeletal muscle and the heart despite the two tissues having different preferred energy sources.

Introduction

Mitochondria, the powerhouse of the cell, produce energy in the form of ATP from the breakdown of carbohydrates, fats, and proteins. Mitochondrial abundance and their preferences for metabolic substrates differ in various cell types 1. While mitochondria are abundant in red skeletal muscles such as Soleus, and the heart, the mitochondrial content in white muscles, such as Extensor Digitorum Longus (EDL), is much lower. Mitochondria in the skeletal muscle and heart have distinct preferences for energy sources - skeletal muscle mitochondria prefer glucose (the major simple carbohydrate) as the substrate, whereas cardiac mitochondria mainly use fatty acids as the fuel. Moreover, different types of skeletal muscle cells (fibers) utilize glucose to produce energy in different mechanisms - in slow-twitch oxidative muscle fibers glucose (via pyruvate) is primarily oxidized in mitochondria for energy production, whereas fast- twitch muscle fibers use cytosolic glycolysis to quickly produce energy to support rapid contraction. Nevertheless, both muscle and heart mitochondria are flexible in selecting substrates in response to stress (exercise, fasting, etc.) and under disease conditions (heart failure, diabetes, etc.) 2–5. It remains to be determined how mitochondrial utilization of various competing metabolic substrates is coordinated and whether disruption of mitochondrial flexibility in fuel selection affects the function of the heart and skeletal muscle.

PTPMT1, encoded by nuclear DNA, is localized to the mitochondrion and anchored at the inner membrane 6. It dephosphorylates phosphatidylinositol phosphates (PIPs). PIPs are a class of membrane phospholipids that bind to a distinctive set of effector proteins, thereby regulating a characteristic suite of cellular processes, including membrane trafficking and ion channel/transporter functions 7,8. PTPMT1 is also involved in the synthesis of cardiolipin by converting phosphotidylglecerol phosphate to phosphotidylglecerol, the precursor of cardiolipin 9. Global knockout of PTPMT1 results in developmental arrest and post-implantation lethality 9,10. Our previous studies suggest that PTPMT1 facilitates mitochondrial metabolism largely by dephosphorylation of downstream PIP substrates 11–13 that appear to inhibit mitochondrial oxidative phosphorylation but enhance cytosolic glycolysis by activation of mitochondrial uncoupling protein 2 (UCP2) 11, a transporter of four-carbon (C4) dicarboxylate intermediates of the tricarboxylic acid (TCA) cycle that has been shown to regulate cellular energetics by limiting mitochondrial oxidation of glucose 14–18. PTPMT1 is highly expressed in the heart and skeletal muscle 10. In the present study, we exploit PTPMT1 tissue-specific knockout models to address the role and mechanisms of PTPMT1-facilitated mitochondrial metabolism in these tissues. We find that PTPMT1 plays an important role in facilitating mitochondrial utilization of carbohydrates and that balanced fuel selection is essential for maintaining both muscle and heart functions despite the two tissues having distinct preferences for energy sources.

Results

Knockout of PTPMT1 from skeletal muscles results in defective contractile function and progressive muscle atrophy

To determine the role of PTPMT1-mediated metabolism in the skeletal muscle and heart, we generated tissue-specific PTPMT1 knockout mice (PTPMT1fl/fl/CKMM-Cre+) by crossing PTPMT1 floxed mice 11 and muscle creatine kinase promoter-driven Cre transgenic mice (CKMM-Cre), which express the Cre recombinase in the skeletal muscle and heart starting at embryonic day 13 19. Quantitative reverse transcription PCR (qRT-PCR) showed ∼95% and ∼80% deletion of PTPMT1 in the skeletal muscle and heart in adult PTPMT1fl/fl/CKMM-Cre+ mice, respectively (Figure 1A). These animals were indistinguishable from their wild-type (WT, PTPMT1+/+/CKMM-Cre+) littermates up to 3 months of age, suggesting that PTPMT1 deletion did not affect muscle or heart development. The knockout mice exhibited reduced weight gain starting at 4 to 5 months (Figure 1A). Muscle wasting was observed in PTPMT1fl/fl/CKMM-Cre+ mice at 8 months or older. Histopathological examination of skeletal muscles of these knockout mice revealed characteristic changes of muscle atrophy - marked variations in muscle fiber diameters and the presence of ghost fibers or more interstitial space (Figure 1B). In addition, Masson’s trichrome staining revealed increased collagen-rich fibrotic regions in PTPMT1 knockout muscles (Figure 1B). We then examined PTPMT1 knockout mice at 6-8 months before muscle wasting occurred. PTPMT1fl/fl/CKMM-Cre+ mice became fatigued faster and fell more quickly than PTPMT1+/+/CKMM-Cre+ control animals in wire hang tests, indicating muscle weakness (Figure 1C). In treadmill exercise tests, both male and female knockout mice displayed reduced maximum speed (Figure 1D), endurance (Figure 1E), and distance of the run (Supplemental Figure 1A). To determine whether the reduced muscle strengths of the knockout mice resulted from muscle cell-intrinsic defects, isometric contractile properties of isolated muscles were examined. Ex vivo force-frequency measurements were performed on Soleus, a more oxidative muscle with a higher percentage of slow-twitch Type I fibers, and EDL, a more glycolytic muscle with a higher percentage of fast-twitch Type II fibers. Both knockout Soleus and EDL showed decreased optimal length, the length of the muscle at which maximal force was achieved (Supplemental Figure 1, C and D). Specific force production of PTPMT1 knockout Soleus (Supplemental Figure 1E) and EDL (Supplemental Figure 1E) was decreased. However, normalized force production was decreased only in knockout Soleus (Figure 1F), but not in knockout EDL (Figure 1G). These data suggest that PTPMT1 plays a more important role in oxidative than glycolytic muscle fibers although both knockout Soleus and EDL developed muscle atrophy subsequently.

Deletion of PTPMT1 from skeletal muscles results in defective contractility and progressive muscle atrophy.

(A) Representative 8-month-old of PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice, and hind limbs and hearts dissected from these mice were photographed. EDL and Soleus (n=8 mice/genotype), and heart (n=4 mice/genotype) weights were measured. PTPMT1 mRNA levels in EDL, Soleus, and heart tissues were determined by quantitative reverse transcription PCR (qRT-PCR) (n=4 mice/genotype). (B) Skeletal muscle sections prepared from 8-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM-Cre+ mice were processed for H&E staining and Masson’s Trichrome staining. One representative image from 3 mice/genotype is shown. (C) Six-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM-Cre+ mice (n=7/genotype) were subjected to wire hang tests. Relative forelimb muscle strength was determined. (D and E) Seven to eight-month-old PTPMT1+/+/CKMM-Cre+ (n=5 males and 5 females) and PTPMT1fl/fl/CKMM-Cre+ (n=3 males and 5 females) mice were assessed by treadmill exercise tests as described in Materials and Methods. Maximum speed (D) and duration of the run (E) were recorded. (F and G) Soleus (F) and EDL (G) dissected from 7 to 8-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM- Cre+ mice (n=3 mice, 6 muscles/genotype) were subjected to ex vivo isometric force measurements. Specific contractile forces produced at the indicated frequencies of stimulation were normalized to the physiological cross-sectional area. Shown on the left are representative absolute forces produced at 1, 10, and 100 Hz. * p < 0.05.

To determine the impact of PTPMT1 depletion from the skeletal muscle and heart on the metabolism of the whole body, we measured plasma glucose and fatty acid levels and found that they were comparable in PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ control mice (Supplemental Figure 2A). Plasma levels of lactate (Supplemental Figure 2B) and lipid metabolic products (Supplemental Figure 2C) in PTPMT1 knockout mice were also relatively normal (except that cholesterol and nonesterified fatty acids levels were decreased and marginally increased, respectively, in female knockout mice). Adipokine array assays of the plasma showed no or subtle differences in Adiponectin, Leptin, Lipocalin- 2, ANGPTL3, Resistin, FGF-21, DPPIV, Fetuin A, IGF-1, etc. between PTPMT1 knockout and control mice (Supplemental Figure 2D). In addition, no differences in glucose tolerance tests were observed between PTPMT1 knockout and control mice (Supplemental Figure 2E). Furthermore, we checked expression levels of key enzymes involved in glucose and lipid metabolism in PTPMT1 knockout muscles. No changes in HK2, PKM1, LDHA, DICER1, MCAD (ACADM), ACADL, HADHA, CPT2, or FABP3 were found. However, the expression of CPT1B, was increased in PTPMT1 knockout muscles (Supplemental Figure 2, F and G). These results suggest that systemic glucose and lipid metabolism in these tissue-specific knockout mice was not significantly changed, but fatty acid partitioning for oxidation in the mitochondria was elevated in PTPMT1 knockout muscles.

PTPMT1 depletion ultimately leads to mitochondrial damage and bioenergetic stress in knockout skeletal muscles

Gömöri trichrome staining revealed ragged red fibers in PTPMT1 knockout skeletal muscles (Figure 2A), as observed in various types of human mitochondrial myopathies. Electron microscopic analyses showed that interfibrillar mitochondria in knockout Soleus and EDL were disorganized and enlarged, in contrast to normal mitochondria that typically reside in pairs and position on either side of the Z-disc in control muscles. In addition, there was a massive accumulation of subsarcolemmal mitochondria in the knockout muscles, whereas the content of intermyofibrillar mitochondria decreased (Figure 2B). Although EDL, unlike Soleus, uses more cytosolic glycolysis than mitochondrial oxidative phosphorylation for energy production, the mitochondrial structural changes appeared to be more severe in EDL than Soleus in the knockout mice (Figure 2B). Furthermore, a massive accumulation of intramyofibrillar lipids were detected by Oil Red O staining in 6-month or older PTPMT1 knockout muscles (Figure 2C).

PTPMT1 loss ultimately leads to abnormal mitochondrial distribution, structural damage, and bioenergetic stress in skeletal muscles.

(A) Skeletal muscle sections prepared from 6-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM-Cre+ mice were processed for Gömöri trichrome staining. One representative image from 3 mice/genotype is shown. (B) Soleus and EDL dissected from 8-month-old PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM- Cre+ mice were processed for transmission electron microscopic examination. One representative image from 3 mice/genotype is shown. (C) Skeletal muscle sections prepared from 6-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM-Cre+ mice were processed for Oil Red O staining to visualize lipids. One representative picture from 3 mice/genotype is shown. (D) Total DNA was extracted from Soleus and EDL dissected from 8-month-old PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice (n=3/genotype). Mitochondrial content was estimated by comparing mtDNA (Cytochrome B) levels to genomic DNA levels by qPCR. (E) Total ATP levels in Soleus and EDL dissected from 8-month-old PTPMT1fl/fl/CKMM- Cre+ and PTPMT1+/+/CKMM-Cre+ mice (n=6/genotype) were determined. (F) Whole cell lysates prepared from the Soleus and EDL isolated from 7-month-old PTPMT1fl/fl/CKMM-Cre+ and their control mice were examined by immunoblotting with the indicated antibodies. Representative results from 3 mice/genotype are shown.

Total mitochondrial content in PTPMT1 knockout Soleus and EDL also slightly decreased compared to that in control muscles (Figure 2D). ATP levels in both knockout Soleus and EDL decreased significantly compared to WT counterparts (Figure 2E). Consistent with these data, the energetic stress sensor, AMP- activated kinase (AMPK), was highly activated in PTPMT1fl/fl/CKMM-Cre+ muscles as determined by the phosphorylation levels of Thr172 (Figure 2F). Acetyl-CoA carboxylase (ACC), a negative regulator of fatty acid oxidation, was inhibited, as evidenced by a marked increase in the inhibitory phosphorylation of this enzyme (Figure 2F), implying increased fatty acid oxidation in PTPMT1-depleted mitochondria. In agreement with previous findings that AMPK negatively regulates mTOR signaling 20,21, mTOR was substantially inhibited. Activities of S6K, S6, and 4E-BP1, key downstream components of mTOR signaling, were also concomitantly decreased (Figure 2F), indicating diminished anabolic activities and consistent with the muscle atrophy phenotype. Interestingly, Akt activities, as reflected by its phosphorylation levels (Ser473 and Thr308), were increased in PTPMT1 knockout skeletal muscles, recapitulating mTOR and raptor (a component of the mTOR complex 1) knockout muscles 22,23. Likely due to the cross-talk between Akt and Ras signaling pathways, Erk activities were also slightly increased in these knockout muscles (Figure 2F).

Given that muscle atrophy developed in PTPMT1fl/fl/CKMM-Cre+ mice after 6-8 months, to assess the direct impact of PTPMT1 deletion on mitochondrial metabolism, it is important to examine mitochondria isolated from young PTPMT1 knockout mice before the tissue damage. Expression levels of electron transport chain complexes in these PTPMT1-ablated mitochondria were not changed (Supplemental Figure 3A). Total cellular ATP levels in PTPMT1 knockout muscles at this age were comparable to those in control tissues at this age (Supplemental Figure 3B), and the knockout muscles did not show bioenergetic/metabolic stress (Supplemental Figure 3C). We also checked LC3-I/II levels in PTPMT1 knockout muscles and found no evidence of elevated autophagic activities (Supplemental Figure 3D). These results indicate that the decrease in total cellular ATP levels in older knockout muscles (Figure 2E) was a secondary effect not a direct effect of PTPMT1 depletion.

PTPMT1 loss impairs mitochondrial utilization of pyruvate, whereas the fatty acid utilization is enhanced

We next sought to address the mechanisms by which PTPMT1 loss impacts mitochondrial metabolism. To avoid secondary effects, we examined mitochondrial metabolic activities of PTPMT1 knockout muscles in 3-month-old PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice. Knockout muscle tissues showed decreased maximal reserve oxidative capacities in the presence of full substrates (Figure 3A), consistent with our previous observations from other PTPMT1 knockout cell types 11,12,24. We also isolated mitochondria from young skeletal muscles and assessed mitochondrial metabolism in the presence of ADP and a single metabolic substrate by real-time measurement of ATP synthesis-driven oxygen consumption. When pyruvate, a key metabolite derived from glucose, was provided as the sole substrate, ATP synthesis-driven oxygen consumption in PTPMT1-deficient mitochondria was markedly decreased (Figure 3B), suggesting that mitochondrial pyruvate utilization was impaired in the absence of PTPMT1. Interestingly, we observed enhanced oxygen consumption in these mitochondria when fatty acids were supplied as the fuel (Figure 3C). In addition, a slightly increased respiratory response in PTPMT1-depleted mitochondria was detected when glutamate, another mitochondrial substrate, was provided (Figure 3D). Importantly, feeding with succinate, the substrate for Complex II of the electron transport chain, resulted in similar oxygen consumption in PTPMT1 null and control mitochondria (Figure 3E), validating the intact function of the electron transport chain in the knockout mitochondria and excluding flavin adenine dinucleotide and quinone-linked deficiencies. These mechanistic data suggest that PTPMT1 plays an important role in facilitating carbohydrate oxidation in mitochondria.

PTPMT1 ablation impairs mitochondrial utilization of pyruvate, whereas the fatty acid utilization is enhanced.

(A) Muscle cross-sections prepared with biopsy punches from PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice (n=4/genotype) at 4 months of age were measured for OCRs at the basal level and following the addition of oligomycin (8 μM), FCCP (4 μM), and antimycin A/rotenone (1 μM). (B-E) Mitochondria were isolated from the skeletal muscles dissected from PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice (n=3/genotype) at 3 months of age. Mitochondrial oxygen consumption (10 μg of mitochondrial protein) was measured in the presence of pyruvate (5 mM)/malate (5 mM) (B), palmitoyl-CoA (40 μM)/carnitine (40 μM)/malate (5 mM) (C), glutamate (5 mM)/malate (5 mM) (D), or succinate (10 mM) (E), following the addition of ADP (4 mM), oligomycin (1.5 μM), FCCP (4 μM), and antimycin A/rotenone (1 μM). Experiments were repeated three times with three independent pairs of mice. Similar results were obtained in each experiment. (F-H) Levels of pyruvate (F), α- KG (G), and acetyl-CoA (H) in the lysates of the mitochondria isolated from the above skeletal muscles were measured (n=6/genotype). (I) Mitochondria freshly isolated from the skeletal muscles of PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice (n=4/genotype) were washed three times in MAS buffer. The mitochondria were then incubated with pyruvate (5 mM)/malate (5 mM) and ADP (4 mM) at 37°C. Five min later, mitochondria were collected, washed, and lysed. α-KG levels in the mitochondrial lysates were measured. (J) Pyruvate dehydrogenase (PDH) activities in the mitochondrial lysates were determined (n=5-6 mice/genotype). (K) MYH4, MYH2, and MYH7 mRNA levels in the skeletal muscles dissected from 3-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM-Cre+ mice (n=4/genotype) were determined by qRT-PCR.

Consistent with the above notion, intramitochondrial pyruvate levels in the mitochondria freshly isolated from PTPMT1 knockout muscles decreased by half (Figure 3F). Levels of α-ketoglutarate (α-KG), an important metabolite in the TCA cycle, also decreased in the knockout mitochondria (Figure 3G), although steady-state mitochondrial acetyl-CoA levels were not changed (Figure 3H). To further determine if pyruvate transport efficiency may be reduced in PTPMT1-ablated mitochondria, we incubated fresh mitochondria with pyruvate and measured acute production of α-KG from extramitochondrial pyruvate. As shown in Figure 3I, α-KG production in PTPMT1-ablated mitochondria was indeed decreased compared to that in control mitochondria. Notably, the activity of pyruvate dehydrogenase (PDH), the enzyme that converts pyruvate to acetyl-CoA to feed the TCA cycle, was not affected in the knockout mitochondria (Figure 3J). These observations together with that the mitochondrial pyruvate carrier/transporter (MPC) was comparably expressed in PTPMT1-ablated mitochondria (Figure 2F) suggest that the function of PTPMT1 null mitochondria in uptaking pyruvate was diminished. Possibly due to an adaptive response to the inhibited pyruvate oxidation in the mitochondria, skeletal muscle fiber-type switching occurred in PTPMT1fl/fl/CKMM-Cre+ mice. Muscle fibers in these knockout mice switched from oxidative to glycolytic fibers, as evidenced by a dramatic increase (∼170-fold) in the levels of MYH4, a marker of glycolytic fast-twitch Type 2B fibers in PTPMT1 knockout mice (Figure 3K).

PTPMT1fl/fl/CKMM-Cre+ mice manifest late onset cardiac dysfunction

The heart in PTPMT1fl/fl/CKMM-Cre+ mice was less impacted than the skeletal muscle although PTPMT1 was ∼80% deleted from the heart (Figure 1A). At 7 months, when skeletal muscles in these mice displayed atrophy and myocyte damage, no evident histological changes were observed in heart tissues (Supplemental Figure 4A). Echocardiographic examination showed no difference in left ventricle (LV) contractility in PTPMT1 knockout mice as reflected by similar LV fractional shortening (FS), Ejection Fraction (EF), and the ratio of peak velocity of early to late filling of mitral inflow (E/A) between knockouts and control mice (Supplemental Figure 4B). Mitochondrial content in PTPMT1fl/fl/CKMM- Cre+ heart tissues was comparable to that in PTPMT1+/+/CKMM-Cre+ control tissues (Supplemental Figure 4C). Total ATP levels in PTPMT1 knockout heart tissues were marginally but not significantly decreased (Supplemental Figure 4D). Moreover, mitochondria in knockout cardiomyocytes showed minimal structural changes (Supplemental Figure 4E). Consistent with these observations, no bioenergetic/metabolic stress was detected in the knockout cardiomyocytes according to the activation status of AMPK and mTOR signaling (Supplemental Figure 4F).

However, severe cardiac dysfunction was observed in 10 to 12-month-old PTPMT1fl/fl/CKMM-Cre+ mice. Histopathological examination revealed that cardiomyocytes in these animals had uneven sizes and were disorganized (Figure 4A). In addition, fibrotic lesions, indicative of myocardial damage, were observed. Echocardiographic examination showed that LV contractility, as measured by FS and EF in M-mode echocardiographic tracing of LV (Figure 4, B-D) and recorded by M-Mode echocardiography movies (Supplemental Video 1, A and B), was markedly decreased in PTPMT1 knockout mice. The E/A ratio determined in pulsed-wave Doppler recording of mitral valve inflow was doubled in the PTPMT1fl/fl/CKMM-Cre+ heart (Figure 4, E and F). Collectively, these data suggest that PTPMT1 depletion also impaired cardiac function albeit later than skeletal muscle dysfunction.

PTPMT1fl/fl/CKMM-Cre+ mice manifest late-onset cardiac dysfunction.

(A) Heart tissue sections prepared from 12-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM- Cre+ mice were processed for H&E staining. One representative image from 3 mice/genotype is shown. (B-F) Cardiac morphology and function of 12-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM-Cre+ mice (n=5/genotype) were examined by echocardiography. Representative long-axis views in M-mode echocardiographic tracing of LV are shown (B). LV FS (C) and LV ejection fraction (EF) (D) were determined. Representative pulsed-wave Doppler recordings of mitral valve inflow are shown (E). Ratios of peak velocity of early to late filling of mitral inflow (E/A) were determined (F).

Heart-specific deletion of PTPMT1 leads to dilated cardiomyopathy and heart failure

Given that PTPMT1 was only ∼80% deleted from the heart in PTPMT1fl/fl/CKMM-Cre+ mice and that impaired skeletal muscle function in these animals may potentially complicate the heart phenotypes, to further investigate the role of PTPMT1-mediated metabolism specifically in cardiac muscles, heart- specific PTPMT1 knockout mice (PTPMT1fl/fl/MYH6-Cre+) were generated by crossing PTPMT1 conditional mice 11 and MYH6-Cre transgenic mice, which express the Cre recombinase specifically in cardiomyocytes 25. These knockout mice appeared healthy in the first several months although PTPMT1 was nearly completely deleted from the heart (but not in Soleus and EDL) (Supplemental Figure 5A). No obvious tissue morphological changes or fibrotic lesions were found in PTPMT1 knockout hearts (Supplemental Figure 5B), confirming that PTPMT1-mediated metabolism is dispensable for the development of the heart. Echocardiography also showed normal heart function in 3-month-old PTPMT1fl/fl/MYH6-Cre+ mice, as evidenced by normal FS, EF, and E/A ratios (Supplemental Figure 5, C- E). Knockout mice did not exhibit any defects during treadmill exercises at 6 months of age (Supplemental Figure 5F). However, all PTPMT1fl/fl/MYH6-Cre+ knockout mice died at 10-16 months, and they often succumbed suddenly (Figure 5A) although their body weights were relatively normal. PTPMT1 knockout hearts showed enlarged ventricular chambers. Severe structural damage in cardiac myocytes and tremendous fibrotic lesions were observed in the knockout hearts (Figure 5B).

Deletion of PTPMT1 from the heart ultimately leads to dilated cardiomyopathy and heart failure.

(A) Kaplan-Meier survival curves of PTPMT1fl/fl/MYH6-Cre+ (n=18) and PTPMT1+/+/MYH6-Cre+ mice (n=20). (B) Heart tissue sections prepared from 10 to 12-month- old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice (n=4/genotype) were processed for H&E staining and Masson’s Trichrome staining. Representative images are shown. (C) Heart tissue sections prepared from 11-month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice were processed for immunofluorescence staining for cleaved caspase 3 followed by DAPI counterstaining. One representative image from 3 mice/genotype is shown. (D-J) Eleven to twelve-month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice (n=5/genotype) were examined by echocardiographic evaluation of ventricular function. Representative long- (upper panel) and short- (lower panel) axis views in M-mode echocardiographic tracing of LV (D), and pulsed-wave Doppler recordings of mitral valve inflow (E) are shown. Left ventricular internal diameters at diastole (LVID-d) and systole (LVID-s) were measured (F). LV ejection fraction (EF) and fractional shortening (FS) (G), ratios of peak velocity of early to late filling of mitral inflow (E/A) (H), the thickness of LV anterior wall at the end of diastole (LVAW-d) and the end of systole (LVAW-s) (I), and thickness of LV posterior wall at the end of diastole (LVPW-d) and at the end of systole (LVPW-s) (J) were determined.

Immunostaining of cleaved caspase 3 illustrated increased apoptosis in PTPMT1fl/fl/MYH6-Cre+ cardiomyocytes as compared to control cells (Figure 5C). Echocardiography revealed a profound dilated cardiomyopathy and heart failure in PTPMT1fl/fl/MYH6-Cre+ mice, including both systolic and diastolic LV dilation, thinning of ventricular walls, depression of EF and FS, arrhythmias, and impaired myocardial contraction (Figure 5, D-J and Supplemental Video 2, A and B). Cardiac strain images demonstrated that the global function of PTPMT1fl/fl/MYH6-Cre+ hearts, as reflected by global longitudinal strain, was significantly decreased (Supplemental Figure 6, A and B), similar to that observed in PTPMT1fl/fl/CKMM-Cre+ mice.

Unlike those in younger knockout mice, total cellular ATP levels in 10 to 12-month-old PTPMT1fl/fl/MYH6-Cre+ hearts were decreased by half (Figure 6A), and the knockout cardiomyocytes exhibited pronounced energetic stress as demonstrated by a substantial increase in p-AMPK (T172) and decrease in p-mTOR (S2448) (Figure 6B). To determine the direct effects of PTPMT1 loss on cardiomyocyte metabolism, we examined heart tissues isolated from young (3 months old) knockout mice before cardiomyocyte and mitochondria damages occurred. The mitochondrial content in these young PTPMT1 knockout cardiac tissues was comparable to that in control tissues (Supplemental Figure 7A), and there were no changes in the expression levels of mitochondrial complexes (Supplemental Figure 7B). Moreover, no bioenergetic/metabolic stress was detected in the knockout cardiomyocytes at this age (Supplemental Figure 7C). However, like PTPMT1 knockout skeletal muscle mitochondria (Figure 3, B-D), PTPMT1-ablated cardiac mitochondria also showed altered fuel selection – the utilization of pyruvate was decreased (Figure 6C), whereas the free fatty acid and glutamate utilization was increased in PTPMT1fl/fl/MYH6-Cre+ cardiac mitochondria (Figure 6, D and E). Intramitochondrial pyruvate levels in PTPMT1-ablated cardiac mitochondria decreased (Figure 6F) while acetyl-CoA levels were not changed (Supplemental Figure 7D), consistent with the metabolite data obtained from PTPMT1- depleted skeletal muscle mitochondria (Figure 3, F and H). Furthermore, acute production of α-KG from extramitochondrial pyruvate was decreased in the cardiac mitochondria lacking PTPMT1 (Figure 6G), although PDH activity in these mitochondria was not significantly affected (Figure 6H).

PTPMT1 deficiency causes mitochondrial substrate shift and oxidative stress, leading to mitochondrial damage and lipid accumulation in PTPMT1 knockout cardiomyocytes.

(A) Total ATP levels in the heart tissues dissected from PTPMT1fl/fl/MYH6- Cre+ and PTPMT1+/+/MYH6-Cre+ mice at the indicated ages (n=5 mice/genotype). (B) Whole cell lysates prepared from the heart tissues of 10 to 12-month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice were examined by immunoblotting with the indicated antibodies. Representative results from 3 mice/genotype are shown. (C-E) Oxygen consumption of the mitochondria isolated from the heart tissues of 2 to 3-month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice (n=3/genotype) was measured in the presence of pyruvate (5 mM)/malate (5 mM) (C), palmitoyl-CoA (40 μM)/carnitine (40 μM)/malate (5 mM) (D), or glutamate (5 mM)/malate (5 mM) (E), following the addition of ADP (4 mM), oligomycin (1.5 μM), FCCP (4 μM), and antimycin A/rotenone (1 μM). OCRs at basal levels, in response to ADP addition, and maximal reserve capabilities were determined. (F) Levels of pyruvate in the lysates of the cardiac mitochondria isolated from 2 to 3-month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice (n=6/genotype) were determined. (G and H) Mitochondria isolated above were washed three times in MAS buffer and then incubated with pyruvate (5 mM)/malate (5 mM) and ADP (4 mM) at 37°C. Five min later, the mitochondria were collected, washed, and lysed. α-KG levels in the mitochondrial lysates were measured (n=4 mice/genotype) (G). Pyruvate dehydrogenase (PDH) activities in the mitochondrial lysates were determined (n=5 mice/genotype) (H). (I) Heart tissue sections prepared from 3-month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice (4 mice/genotype) were processed for ROS staining. One representative picture from 4 mice/genotype is shown. (J) Heart tissues dissected from 6 to 8- month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice were processed for TEM examination. One representative image from 4 mice/genotype is shown. (K) Heart tissue sections prepared from 10 to 12-month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice were processed for Oil Red O staining to visualize lipids. One representative picture from 3 mice/genotype is shown.

Oxidation of fatty acids in mitochondria yields much more ATP relative to carbohydrates (via pyruvate); however, fatty acid oxidation requires a greater rate of oxygen consumption for a given rate of ATP synthesis than carbohydrates. Such that the byproduct reactive oxygen species are increased when mitochondria switch to fatty acids for energy metabolism. Indeed, PTPMT1 knockout heart tissues showed elevated overall cellular ROS levels even at 3 months (Figure 6I). Persistent oxidative stress can cause mitochondrial damage. Electron microscopic examination revealed that the myocardium of PTPMT1fl/fl/MYH6-Cre+ mice was distinguishable from that of PTPMT1+/+/MYH6-Cre+ control mice by the structures of mitochondria after 6-8 months (Figure 6J). Similar to that observed in PTPMT1 knockout skeletal muscle cells (Figure 2C), drastic accumulation of lipids or lipid intermediates was observed in 10-12-month-old PTPMT1fl/fl/MYH6-Cre+ cardiomyocytes (Figure 6K), which is likely responsible for the increased apoptosis and fibrillation in the knockout heart (Figure 5, B and C) due to the toxicity of excess lipids (lipotoxicity) 26,27.

PTPMT1 is dispensable for the liver and adipose tissues

Finally, to further determine whether PTPMT1-facilitated carbohydrate oxidation in mitochondria is also important for other important metabolic tissues, we generated liver-specific (PTPMT1fl/fl/ALB-Cre+) and fat-specific (PTPMT1fl/fl/Adipoq-Cre+) PTPMT1 knockout mice by crossing PTPMT1 floxed mice 11 with ALB-Cre 28 and Adipoq-Cre 29 mice, which express Cre specifically in hepatocytes and adipocytes, respectively. In contrast to muscle/heart- and heart-specific knockout mice, liver-specific PTPMT1 knockout mice were healthy without overt abnormalities. They had a normal lifespan. Their body weights, liver weights, and blood glucose levels under both feeding and fasting conditions were comparable to those in control animals (Supplemental Figure 8, A-D). Moreover, no significant differences were observed in glucose tolerance tests between PTPMT1fl/fl/ALB-Cre+ and control mice (Supplemental Figure 8E). Unlike muscle/heart- and heart-specific PTPMT1 knockout mice in which prominent lipid accumulation was detected in muscle and cardiac myocytes, liver-specific PTPMT1 knockout mice did not display this secondary effect in hepatocytes, and there was no evident tissue damage in the liver or a fatty liver phenotype (Supplemental Figure 8F). Similarly, adipocyte-specific PTPMT1 knockout mice were also healthy. Their body weights and glucose tolerance were comparable to those of control animals (Supplemental Figure 9, A and B). No histological changes were observed in PTPMT1 knockout white adipose and brown adipose tissues (Supplemental Figure 9C). The fact that PTPMT1fl/fl/ALB-Cre+ and PTPMT1fl/fl/Adipoq-Cre+ mice well tolerated PTPMT1 deficiency in hepatocytes or adipocytes suggests that the detrimental effects of PTPMT1 depletion-induced mitochondrial substrate shift are highly cell type-specific.

Discussion

Mitochondrial flexibility in substrate selection is essential for maintaining cellular energy homeostasis in physiology and under stress; however, the regulatory mechanisms and its significance for various tissues and cell types remain poorly characterized 1. In this study, we have identified an important role of the mitochondrial phosphatase PTPMT1 in maintaining mitochondrial flexibility. The loss of PTPMT1 decreased mitochondrial utilization of pyruvate, a key mitochondrial substrate derived from glucose, whereas fatty acid oxidation was enhanced (Figure 3C, Figure 6D). Importantly, this prolonged mitochondrial substrate shift causes oxidative stress and mitochondrial damage, ultimately leading to the accumulation of lipids/lipid intermediates in PTPMT1 knockout skeletal muscle and cardiac myocytes. Excess lipids/lipid intermediates are known to be toxic to cardiomyocytes (lipotoxicity) 26,27. The profound damages in PTPMT1 knockout hearts and muscles that occurred subsequently were likely attributed to these accumulated lipids, although other factors, such as decreased anabolic activities (as demonstrated by diminished mTOR signaling) (Figure 2F, Figure 6B), could also contribute to these phenotypes. The mitochondrial and tissue damages combined with the resulting energetic stress finally led to muscle atrophy and heart failure in the knockout mice.

One of the important findings in this report is that persistent mitochondrial fuel switch from pyruvate to fatty acids is detrimental for both cardiac and skeletal muscles despite the two tissues having different preferred energy sources. Glucose is the major energy source for skeletal muscles. It is used for energy production mainly in mitochondria in oxidative red muscles or through cytosolic glycolysis in glycolytic white muscles. Glucose is also the main energy source for the developing heart, but the adult heart primarily utilizes fatty acids to produce ATP. Indeed, when mitochondrial utilization of pyruvate was inhibited by PTPMT1 deletion, heart dysfunction developed much later than skeletal muscle dysfunction. Interestingly, our data from these animal models suggest that efficient carbohydrate oxidation in mitochondria and mitochondrial balanced fuel selection are critically important for both oxidative and glycolytic muscles as well as the adult heart, but not for the development of the skeletal muscle and heart. Despite enhanced fatty acid and glutamate oxidation and that energy homeostasis was initially maintained in PTPMT1 knockout muscle cells and cardiomyocytes due to metabolic adaptations or activation of alternative energy-producing pathways, these knockout cells later still showed energy deficits due to oxidative stress and mitochondrial damage that occurred subsequently. PTPMT1 knockout hepatocytes or adipocytes did not display such secondary effects (mitochondria and cell damages). This is possibly because the requirement of carbohydrate oxidation for energy production in hepatocytes and adipocytes is lower than that in skeletal muscle cells and cardiomyocytes, and/or PTPMT1 knockout hepatocytes and adipocytes did not need to uptake excessive fatty acids in the effort to maintain energy homeostasis. Another possibility is that the capabilities of hepatocytes and adipocytes to process fatty acids and to clear lipid metabolites (through cholesterol synthesis and cholesterol exportation out of the cell) are much stronger than muscle cells and cardiomyocytes 30.

Another important finding in this report is that PTPMT1 plays a major role in governing the metabolic fate of pyruvate and maintaining mitochondrial flexible substrate utilization. Pyruvate is at the central branch point of mitochondrial oxidation and cytosolic glycolysis in glucose metabolism − It can enter the mitochondrion for oxidation or be reduced to lactate in the cytosol. Our data suggest that inhibited mitochondrial utilization of pyruvate is most likely responsible for the remodeled metabolism in PTPMT1 knockout cells since ATP synthesis-driven oxygen consumption of PTPMT1-ablated mitochondria was greatly reduced with pyruvate as the sole substrate (Figure 3B, Figure 6C). Moreover, acute production of α-KG from extramitochondrial pyruvate was indeed decreased (Figure 3I, Figure 6G).

The precise mechanism by which PTPMT1 facilitates mitochondrial utilization/uptake of pyruvate remains to be further determined. PTPMT1 is localized to the inner mitochondrial membrane where pyruvate transporters reside. This phosphatase dephosphorylates phosphatidylinositol phosphates (PIPs)11–13. We have shown previously that PIPs directly promote fatty acid-induced activation of UCP2 11. UCP2, unlike UCP1, is a C4 metabolite transporter 14. Elevated UCP2 activity inhibits mitochondrial oxidation of glucose but enhances cytosolic glycolysis although the underlying mechanisms remain unclear 15–18. PTPMT1 depletion likely leads to the buildup of downstream PIP substrates in the mitochondrial inner membrane, which in turn remodel mitochondrial energy source selection by enhancing UCP2 function 11 [UCP2 levels in PTPMT1-ablated mitochondria were not changed (Supplemental Figure 3A, 7B)]. It may also be possible that the accumulated PIPs directly inhibit the mitochondrial pyruvate transporter MPC. It is important to note that PTPMT1 was reported to be involved in the synthesis of cardiolipin 9 which is one of the major components of the mitochondrial inner membrane and that plays an important role in maintaining mitochondrial membrane stability and dynamics. Thus, it may also be possible that PTPMT1 depletion decreases mitochondrial pyruvate utilization through reduced cardiolipin synthesis. Comprehensive metabolomic analyses for PTPMT1-deleted cells would provide additional insights into the mechanisms underlying the metabolic alterations caused by PTPMT1 loss.

In summary, in this report, we provide evidence that the mitochondria-based multi-functional phosphatase PTPMT1 plays an important role in facilitating mitochondrial utilization of pyruvate and maintaining mitochondrial flexibility for balanced substrate selection. Prolonged mitochondrial energy sources switch from carbohydrates to lipids such as that caused by PTPMT1 depletion is detrimental for the skeletal muscle and heart but not the liver and adipose, ultimately leading to muscle atrophy and heart failure.

Materials and methods

Mice

PTPMT1fl/+ mice with the C57Bl/6 background were generated in our previous study 11. CKMM-Cre+ 19 (Stock# 006475), MYH6-Cre+ 25 (Stock# 011038), ALB-Cre+ 28 (Stock# 003574), and Adipoq-Cre+ 29 (Stock# 010803) mice with the C57BL/6 background were purchased from the Jackson Laboratory. To avoid potential side effects of Cre, PTPMT1+/+/Cre+ mice were used as controls for PTPMT1fl/fl/Cre+ mice and the Cre transgene was hemizygous in both mouse types. In initial pilot experiments, we examined PTPMT1+/+/Cre+ mice and no defects were observed in these animals. Mice were kept under specific pathogen-free conditions in Case Western Reserve University Animal Resources Center and Emory University Division of Animal Resources. Mice of the same age, sex, and genotype were randomly grouped for subsequent analyses (investigators were not blinded). All animal procedures complied with the NIH Guidelines for the Care and Use of Laboratory Animals and were approved by the Institutional Animal Care and Use Committee.

Antibodies

The sources and catalog numbers of the antibodies utilized in this study are listed below.

Wire hang test

Maximal muscular strength of mouse forelimbs was assessed by wire hang tests following a standard protocol. Mice were allowed to grasp and hang on a wire with forelimbs. Time duration when the mouse was able to suspend before falling was recorded. Three trials were carried out for each mouse. The average time was normalized against body weight. Values of control mice were set to 100%.

Treadmill exercise test

Treadmill exercise tests were conducted following the protocol previously reported by trained personnel at Case Western Reserve University Mouse Metabolic Phenotyping Center Analytical Core 31,32. Adjustable parameters on the treadmill include belt speed (0-99 m/min) and angle of inclination (0-45”). After acclimation for 1 hour, 2.5 m/min incremental increases in treadmill belt speed and 2’ increments in an angle of inclination were performed every 3 min until the mouse exhibited signs of exhaustion. Exhaustion was defined as the mouse spending >50% of the time or X5 sec consecutively on the shock grid. The maximum speed and duration of the run were recorded at the time of exhaustion.

Ex vivo muscle contractility assay

Intact Soleus and EDL isolated from mice were used for the collection of isometric force data using an eight-chamber system 12,33. Data were analyzed with ADInstruments PowerLab software customized for these experiments. Muscles were first equilibrated for 20 min to mimic conditions of normal activity (low duty cycle, ∼1%). Muscles were then subjected to the length-force relationship test to determine the optimal length at which maximal force is achieved. Muscles were stimulated with frequencies ranging from 1 to 130 Hz to generate force versus frequency relationships. From these relationships, the frequency producing maximal tetanic force (Tmax) was then used for the rest of the experiments. Muscles were stimulated every minute with Tmax for 5–10 min to assure that the preparation was stable at which point the tissue bathing buffer was replaced with one containing 0 mM Ca2+ + 0.1 mM EGTA. The muscles were then stimulated every minute at Tmax to determine the impact of the removal of external calcium on muscle force. After 20 min, the 0 mM Ca2+ buffer was washed out and replaced by a buffer containing 2.5 mM Ca2+, and the muscles were stimulated every minute for 20 min at Tmax to allow for force recovery. After force measurement protocols, muscle dimensions and masses were measured for the determination of cross-sectional normalized forces (relative force). Muscle force was reported as absolute force (mN) and force normalized to the following muscle physiological cross-sectional area (PCSA, N/cm2) equation as previously reported 34,35 with a small modification in that in our final length calculation, we considered the length to weight ratio of muscle fibers for stretched muscles (tendon-to- tendon) of ex vivo muscles in contractility chambers 36 as the actual muscle fiber size was smaller than the length of the measured stretched muscle. Since the PCSA is very sensitive to length, by adjusting to the length of the muscle fiber, force estimation was more precise. Thus, our final formula was: Muscle force (N/cm2) = (force (g) x (muscle length (cm) x 0.75 (EDL) or 0.85 (Soleus) x 1.06 (muscle density))/(muscle weight (g) x 0.00981).

Echocardiography

Echocardiographic studies were performed by trained investigators at Emory University Department of Pediatrics Animal Physiology Core as previously reported 37,38. Mice were anesthetized with 1-2% isoflurane/100% oxygen continuously and positioned on a warming platform set to 37°C. The heart rate was controlled at 360-460 beats per minute to standardize recordings. ECG and respiratory rate were monitored for physiological assessment while scanning. A Vevo 2100 digital high-frequency ultrasound system (FujiFilm Visualsonics Inc, Toronto, ON, Canada) equipped with an MS400, 30-MHz transducer was used to acquire the images. Interventricular septal thicknesses (IVS), left ventricular internal dimensions (LVID) and posterior wall thicknesses (LVPW) at diastole and systole (IVS-d, LVID-d, LVPW-d and IVS-s, LVID-s, LVPW-s) were measured from M-mode images in both parasternal long axis and short-axis views. Left ventricle volume at diastole and systole (LV Vol-d, LV Vol-s), left ventricular ejection fraction (LVEF) and left ventricular fractional shortening (LVFS) were calculated as the following formula (Vevo 2100 Workstation Software VisualSonics INC, Toronto, ON, Canada): LV Vol- d(ul) = [(7.0 /(2.4 + LVID-d)] * LVID-d3); LV Vol-d(ul) = [(7.0 /(2.4 + LVID-s)] * LVID-s3); EF (%) = 100 * [(LV Vol-d – LV Vol-s) / LV Vol-d]; FS (%) = 100 * [(LVID-d – LVID-s) / LVID-d]. The mitral valve flow Doppler velocities were measured by the apical four-chamber view. Early (E) and late (A) mitral valve inflow velocity and mitral valve E to A ratios (MV E/A) were also obtained.

Echocardiographic speckle-tracking based strain measurement was also performed using the VevoStrainTM application (VisualSonics Vevo 2100 Imaging System, FujiFilm VisualSonics, Inc, Toronto, ON, Canada) which produces velocity strain and time-to-peak analyses on myocardial wall images. The B-Mode cine loop was acquired for Vevo Strain analysis. The qualified consecutive cardiac cycles were selected between one respiration cycle and the next in the EKG panel. Both endocardium and epicardium were traced simultaneously and semi-automatically at the parasternal long-axis view. VevoStrain builds the dynamic LV wall trace for all frames in the cine loop. Velocity, displacement, strain, and strain rate data were calculated based on a speckling-tracking algorithm provided by the Vevo software. Data on overall cardiac tissue performance, segmental time-to-peak, and segmental phase quantification were also collected.

Intracellular ATP quantification

Skeletal muscle and heart tissues were lysed in lysis buffer [Tris-HCl (50 mM), pH 7.4, NP-40 (1%), Na- deoxycholate (0.25%), NaCl (150 mM), EDTA (1 mM), NaF (1 mM), Na3VO4 (1 mM), PMSF (1 mM), and Aprotinin/Leupeptin (10 µg/ml)] for 30 min. Samples were then centrifuged (10,000g) for 10 min, and the supernatants were collected. Total ATP levels were assessed using an ATP bioluminescence assay kit (Sigma, St. Louis, MO), following the instructions provided by the manufacturer. ATP levels were normalized against tissue weights or protein contents in the samples.

Measurement of respiration in isolated mitochondria

Mitochondria were freshly isolated from skeletal muscle and heart tissues. Respiration of mitochondria was measured using a Seahorse XF24 analyzer as described previously 39. Mitochondria (10 µg of protein) were plated in each well of a XF24 plate in 50 μl 1x Mitochondrial Assay Solution (MAS), pH 7.4 [Sucrose (70 mM), Mannitol (220 mM), KH2PO4 (5 mM), MgCl2 (5 mM), HEPES (2 mM), EGTA (1 mM), and fatty acids-free BSA (0.2%)]. XF24 plates were centrifuged at 4°C for 20 min at 2,000g, and then 450 μl of MAS containing 5 mM pyruvate/5 mM malate, 5 mM glutamate/5 mM malate, 40 μM palmitoyl-CoA/40 μM carnitine/5 mM malate, or 10 mM succinate was added to each well. Plates were incubated at 37°C in an incubator for 8–10 min. Oxygen consumption rates (OCRs) were measured under basal conditions in the presence of ADP (4 mM), the ATP synthase inhibitor oligomycin (1.5 μM), the mitochondrial uncoupling compound carbonylcyanide-4-trifluorometh-oxyphenylhydrazone (FCCP, 4 μM), and the electron transport chain inhibitor rotenone (1 μM).

Measurement of respiration in fresh muscle tissues

OCRs of muscle tissues were measured using a Seahorse XF24 analyzer and Seahorse XF24 Islet Capture Microplates (Agilent, 101122-100) as described previously 40. Briefly, tibialis anterior muscles were dissected into 3 x 1 mm cross-sections with a biopsy punch (3 mm diameter). The tissue sections were incubated for 1 hour in DMEM supplemented with 10 mM glucose, 2 mM L-glutamine, and 1 mM sodium pyruvate. Tissue OCRs were measured at the basal level and following the addition of oligomycin (8 μM), FCCP (4 μM), and antimycin A/rotenone (1 μM). Following OCR measurements, tissues were dried at 60 °C for 48 hours. All tissue OCR measurements were then normalized against dry tissue weights.

Mitochondrial pyruvate dehydrogenase (PDH) activity assay, pyruvate, acetyl-CoA, and α-ketoglutarate (α-KG) measurements

Mitochondria were freshly isolated from skeletal muscle/heart tissues and washed twice with MAS. Mitochondrial pellets were lysed in [Tris-HCl (pH 7.4, 50 mM), NP-40 (1%), Na-deoxycholate (0.25%), NaCl (150 mM), EDTA (1 mM), NaF (1 mM), Na3VO4 (1 mM), PMSF (1 mM), aprotinin/leupeptin (10 µg/ml)]. Lysates were subjected to the PDH activity assay and pyruvate, acetyl-CoA, and α-KG measurements using PDH activity colorimetric assay, pyruvate assay, acetyl-CoA assay, and α-KG assay kits purchased from BioVision following the manufacturer’s instructions.

Histological and transmission electron microscopic (TEM) examination

Freshly dissected skeletal muscle and heart tissues were embedded in Optimal Cutting Temperature (OCT) compound (Tissue-Tek) and frozen at -80°C. Slides were stained for Hematoxylin and Eosin (H&E), Masson’s trichrome, and Oil Red O. For TEM analyses, mice were perfused with 3% paraformaldehyde, 1.5% glutaraldehyde, 100 mM cacodylate (pH 7.4), and 2.5% sucrose. Skeletal muscle and heart tissues were postfixed for 1 hour and then processed for TEM examination.

Tissue reactive oxygen species (ROS) determination

Freshly dissected heart tissues were embedded in Optimal Cutting Temperature (OCT) compound (Tissue-Tek) and immediately placed on crushed dry ice. Frozen sections were prepared at 8 μm thickness. After being rinsed in pure H2O for 30 seconds to wash out OCT compound, slides were stained with 5 μM dihydroethidium (DHE) at room temperature in dark for 15 min. Slides were washed with deionized H2O for 1 min three times and visualized immediately using a fluorescence microscope. DHE- derived 2-OH-E+, a specific adduct of cellular superoxide, was visualized with a red excitation filter. Equal exposure times were applied to all slides. Fluorescence intensities were measured by ImageJ software.

Immunoblotting

Skeletal muscle and heart tissues were lysed in RIPA buffer (50 mM Tris-HCl pH 7.4, 1% NP-40, 0.25% Na-deoxycholate, 150 mM NaCl, 1 mM EDTA, 1 mM NaF, 1 mM Na3VO4, 10 μg/mL leupeptin, 10 μg/mL aprotinin, and 1 mM PMSF). Lysates (50 μg) were resolved by SDS-PAGE followed by immunoblotting with the indicated antibodies following standard procedures.

Statistics

Data are presented as mean ± S.D. of all mice analyzed (i.e., biological replicates) in multiple experiments. Tissues, mitochondria, etc. that were analyzed were isolated from single individual mice (not pooled). Statistical significance was determined using unpaired two-tailed Student’s t test. * p < 0.05; ** p < 0.01; *** p < 0.001. N.S. indicates not significant.

Supporting information

eLife Assessment

This paper provides a useful set of data examining the role of PTPMT1, a mitochondria-based phosphatase, in mitochondrial fuel selection. The data were collected and analyzed using solid methodology and can be used as a starting point for further studies that build on the findings here.

Reviewer #1 (Public Review):

The manuscript entitled, "Loss of PTPMT1 limits mitochondrial utilization of carbohydrates and leads to muscle atrophy and heart failure," by Zheng, et al., is focused on assessing the role of deletion of PTPMT1, a mitochondria-based phosphatase, in mitochondrial fuel selection. Authors show that the utilization of pyruvate, a key mitochondrial substrate derived from glucose, is inhibited, whereas fatty acid utilization is enhanced. Importantly, while the deletion of PTPMT1 does not impact development of skeletal muscle or heart, the metabolic inflexibility leads to muscular atrophy, heart failure, and sudden death. Mechanistically, authors claim that the prolonged substrate shift from carbohydrates to lipids causes oxidative stress and mitochondrial dysfunction, leading to accumulation of lipids and muscle cell and CM damage in the KO. Interestingly, PTPMT1 deletion from the liver or adipose tissue does not generate any local or systemic defects. Authors conclude that PTPMT1 plays an important role in maintaining mitochondrial flexibility and that the balanced utilization of carbohydrates and lipids is essential for skeletal muscle and heart. While interesting and authors did a ton of experiments for this project, several major concerns exist. First, because both the CKMM- and the MYHC-Cre express early, during development , it seems the effects of the deletion of PTPMT1 are more likely be specific to defects in muscle and cardiac development rather than postnatal, especially since loss of PTPMT1 affects mTOR activity; indeed, previous reports have shown that selective deletion of mTOR or raptor in skeletal muscle during embryonic development leads to a reduction in postnatal growth and the development of late-onset myopathy and premature death around 6 to 8 months of age. The effects of the deletion in muscle seem eerily similar and therefore likely mechanistically function the same -embryonically, as has been previously suggested. This is also true for cardiac abnormalities, where developmental defects can manifest in mice as they age after at least 3-4 months and decreased mTOR activity can lead to significant cardiac dysfunction and failure (similarly to the effects observed here by the authors). To prove one way or another, authors should show developmental data providing evidence that the effects are not occurring at this stage. It is a lot of work, but the right way to differentiate pre- vs post- development functions of PTPMT1 in the muscle and heart, otherwise cannot verify mechanistically what the precise cause for the phenotype may be. Authors could consider generating mice that have inducible Cre drivers. In addition, how is it that the effects of loss of PTPMT1 are similar between muscle and heart given the differences in energy usage and utilization between these two tissues? Increases in AMPK are usually associated with better metabolic outcomes, particularly in the heart. Increased AMPK activation has also been shown to help reduce fat storage, increase insulin sensitivity, reduce cholesterol/triglyceride production, and suppress chronic inflammation. In addition, increases in carnitines are associated with enhanced metabolic function. Carnitines facilitate transport of long-chain fatty acids into the mitochondrial matrix, triggering cardioprotective effects through reduced oxidative stress, inflammation and necrosis of cardiac myocytes. All of these factors are positive, so how do authors explain this discrepancy in their findings which suggest opposing outcomes- as above, I suggest the explanation is that it is due to developmental effects of deletion of PTPMT1.

Authors attribute much of the pathology in the muscle and heart due to increased lipid accumulation in these tissues; but how do authors explain how hearts and muscle have more fat when the mice are smaller than wt? Is there a difference in energy expenditure in the mice? What about changes in white fat, core temperature or browning of fat? Authors do not mechanistically prove that lipid accumulation is the cause of death in these animals. Rescue experiments should be considered.

Reviewer #2 (Public Review):

Zheng et al. have investigated the effects of PTPMT1 Knock-out on cellular metabolic flexibility. Using several types of appropriate tissue-specific mouse models, the authors have generated data that are both reasonable and broadly significant. While the central mechanism driving the metabolic fuel preference and flexibility remains elusive as the author mentioned in the main text, the finding that the absence of PTPMT1 inhibits glucose (pyruvate) utilization and promotes FAO, resulting in cellular stress and damage, particularly in skeletal and cardiac muscle cells, is intriguing and has practical implications for further research. However, some quantitative data are lacking and certain explanations may be misleading, warranting revisions.

Author Response

The following is the authors’ response to the current reviews.

Comment 1: The descriptions about body weights should be matched.

Regrettably, we did not monitor the body weights throughout the study. We have now revised the description clarifying the confusions. Importantly we evaluated the weights of the muscle (EDL and soleus) and heart tissues in 8-month-old mice (Fig. 1A).

Comment 2: Quantitative data for figures.

As stated in the manuscript, the presented images are representatives of at least three mice per genotype. However, assessing specific measurements such as cell sizes, diameters, or mitochondria sizes in histological tissue sections and electron microscopical fields is not feasible due to practical limitations. Unfortunately, we do not have access to specialized software for such analyses. While semi-quantification of Western blot bands is possible, implementing this for all Western blots in the manuscript would result in a substantial increase in the number of bar graphics. Below are Western blots from additional two pairs of mice used in all figures.

Comment 3: Confusions about “total mitochondrial content”.

The mitochondria content in cells was assessed by quantitatively comparing the DNA level of the mitochondrial gene cytochrome B to that of the nuclear gene 18S using quantitative PCR. This method is commonly used to determine the relative number of mitochondria in cells. However, we have revised and provided a clearer description in the figure legend to avoid any potential confusion.

Comment 4: Suggestions on further analyses of PGC1-alpha and TFAM. LC3-I and -II.

We evaluated LC3-I/II levels in PTPMT1 knockout muscles, and our findings did not indicate any signs of increased autophagic activity (Supplementary Figure S3). We will examine PGC1-alph and TFAM levels in our future studies. It is worth noting that in our previous RNA-seq analyses of PTPMT1 knockout hematopoietic cells, we did not observe any significant alterations in the expression levels of these two genes.

Comment 5: Description on fibrotic lesions.

Quantifying fibrotic areas poses a significant challenge. Therefore, we were only able to describe this finding.

Comment 6: Fig 6 is not well organized and aligned.

In response to your suggestion, we have reorganized this figure accordingly. Panels C, D, and E display mitochondrial OCR data derived from three biological replicates/genotype. We feel that these changes are sufficient to demonstrate the differences in substrate utilization between PTPMT1 knockout and control mitochondria.

Comment 7: Descriptions on glucose oxidation and glycolysis in different types of muscle fibers are confusing

We have followed the suggestions and revised the descriptions accordingly.

Comment 8: A discussion about lactate utilization in cardiomyocytes would be helpful.

Following this suggestion, we have now added a brief discussion.

Comment 9: “Cropped” images were used in Fig 10.

The images shown in Fig. 10 were not cropped images. In order to efficiently use the tissue and mitochondrial lysates, the Western blot membranes were intentionally cut into smaller fragments based on the molecular weights of the proteins to be detected. These smaller membrane sections were then employed for individual Western blotting purposes.

Minor comment 1: The order of Fig 1 panels should be reorganized.

Following this suggestion, we have now reorganized this figure.

Minor comment 2: Suggestion for an Echocardiograph result table.

These analyses were carried out by trained personnel at the Emory Animal Physiology Core. The data presented in our manuscript was provided by them. It is important to note that no additional parameters were measured beyond the data provided by the Core.

Minor comment 3: Is ROS production increased in PTPMT1 knockout muscle cells?

Yes, PTPMT1 knockout tissues showed elevated overall cellular ROS levels even at 3 months (Figure 6I).

Minor comment 4: Typo in S10 legend.

The typo has been corrected.

The following is the authors’ response to the original reviews.

Comment 1: The effects of PTPMT1 on the skeletal muscle and heart might be an embryonic defect. They might be mediated by significantly reduced mTOR signaling

We acknowledge the valid point made by this reviewer. While both CKMM-Cre and Myh6Cre express Cre during the embryonic stage, we did not observe any developmental defects in skeletal muscle-specific (PTPMT1fl/fl/CKMM-Cre) or heart-specific (PTPMT1fl/fl/Myh6-Cre) knockout mice. These knockout mice appeared indistinguishable from their WT littermates until the age of 3-4 months.

Morphologically, the skeletal muscle and heart dissected from these mice showed no abnormalities. Additionally, mitochondria isolated from these tissues did not exhibit any morphological/structural defects. Undoubtedly, the late-onset phenotypes observed in the knockout mice over time was attributed to the metabolic defects arising from the loss of PTPMT1 in the embryos. Although PTPMT1 knockout muscle cells and cardiomyocytes initially maintained energy homeostasis through enhanced fatty acid and glutamate oxidation, along with metabolic adaptations or activation of alternative energy-producing pathways in the first few months, they eventually encountered substantial energy deficits. This was attributed to the subsequent occurrence of oxidative stress and mitochondrial damage. In response to this valuable feedback, we have included a brief discussion in the manuscript's discussion section to address this point.

As mentioned in the manuscript, the late-onset phenotypes observed in our study were likely a result of subsequent damages induced by prolonged metabolic substrate shift and lipid accumulation within the cells. We agree with the reviewer that decreased mTOR activities may also contribute to these late effects, and have included a brief discussion in the discussion section.

Comment 2: Why are the effects of the loss of PTPMT1 similar in the skeletal muscle and heart.

The depletion of PTPMT1 yields similar effects in both tissue types; however, the manifestations occur earlier in the skeletal muscle. Although mitochondria in the skeletal muscle and heart have distinct preferences for energy sources, prolonged forced utilization of fatty acids caused by PTPMT1 depletion eventually leads to lipid accumulation and cellular damage (lipotoxicity) in both tissue types. This phenomenon underscores the importance of maintaining a balance in substrate utilization to prevent adverse effects on cellular health in the skeletal muscle and heart.

Comment 3: AMPK is activated in PTPMT1 knockout cardiomyocytes; this should have cardioprotective effects.

AMPK can be activated through various mechanisms. In our study, AMPK activation occurs in response to energetic stress in late-stage PTPMT1 knockout tissues that displayed significantly reduced ATP levels, aligning with its role as a bioenergetic stress sensor. It is possible that AMPK activation alone was insufficient to overcome the secondary damages induced by the prolonged metabolic switch from carbohydrate metabolism to fatty acid metabolism.

Comment 4: Knockout skeletal muscles and hearts had lipid accumulation; why were knockout mice smaller than controls? Are there any changes in white fat, core temperature or browning of fat? Rescue experiments should be considered to prove that lipid accumulation is the cause of death in the knockout mice.

We believe that the lipid accumulation observed in muscle cells and cardiomyocytes of the knockout mice does not necessarily imply that these tissue-specific knockout mice would be heavier or have increased body fat. We appreciate the suggestions regarding energy expenditure tests and rescue experiments. We will certainly consider incorporating these experiments into our future study.

As stated in the manuscript, we did not observe any morphological changes in white or brown fat tissues in the adipocyte-specific PTPMT1 knockout mice. Furthermore, we assessed body temperature and its response to a cold environment (4°C), and no differences were detected between the knockout mice and the control mice.

Comment 5: Are there sex differences in muscle and heart phenotypes in the tissue specific knockout mice?

We did not observe significant differences in phenotypes between male and female knockout mice.

Comment 6: What happens to UCP2 activity in PTPMT1 deleted cells and what is its function in mediating AMPK and/mTOR regulation.

Currently, there is a lack of direct methods available to measure UCP2 activity. The relationship between UCP2 and the regulation of AMPK and mTOR has not been extensively investigated.

Comment 7: What is the effect of PTPMT1 deletion on cardiolipin synthesis?

PTPMT1 has been implicated in both facilitating mitochondrial utilization of pyruvate and participating in the synthesis of cardiolipin. To investigate the impact of PTPMT1 knockout on cardiolipin levels, we plan to establish a mass spectrometry assay for the quantitative analysis of cardiolipin in knockout mitochondria. Completing these experiments might require a considerable amount of time. Nonetheless, we extensively addressed this point in the discussion section.

Minor concerns:

Comment 8: The title needs more specificity.

As suggested, we have revised the title to "Loss of PTPMT1 restricts mitochondrial utilization of carbohydrates and induces muscle atrophy and heart failure in tissue-specific knockout mice".

Comment 9: Heart and skeletal muscle weights in Fig 1A should be normalized against tibia length.

Unfortunately, we did not perform normalization in this study. However, we appreciate the suggestion and will incorporate it into our future studies. It is important to note that the lengths of tibias in the knockout mice were only marginally shorter.

Comment 10: Low magnification and longitudinal section of the muscle should be shown in Fig 1B and 2A.

The histological images provide supporting evidence for the conclusion, despite not being optimal in quality. We acknowledge the suggested improvements and assure you that we will integrate them into our future studies. It is crucial to emphasize that each conclusion in this study was derived from multiple experimental designs, rather than solely relying on morphological changes.

Comment 11: Fig 1F is mislabeled as 1G.

We have conducted a thorough review and can confidently confirm that the labeling is correct.

Comment 12: Fig 2F and 6B should be quantified.

As indicated in the manuscript, the images presented are representatives of at least three mice per genotype. While semi-quantification of Western blot bands is possible, implementing this for all Western blots in the manuscript would result in a substantial increase in the number of bar graphics. Below are Western blot images from additional two pairs of mice included in Fig. 2F and Fig. 6B. Furthermore, Western blot images from two additional pairs of mice in other figures are also provided below.

Western blotting data from additional two pairs of mice in Fig.

2F.

Western blotting data from additional two pairs of mice in Fig.

6B.

Western blotting data from additional two pairs of mice in Supplementary Fig.

2G.

Western blotting data from additional two pairs of mice in Supplementary Fig.

3A.

Western blotting data from additional two pairs of mice in Supplementary Fig.

3C.

Western blotting data from additional two pairs of mice in Supplementary Fig.

3D.

Western blotting data from additional two pairs of mice in Supplementary Fig.

4F.

Western blotting data from additional two pairs of mice in

Western blotting data from additional two pairs of mice in Supplementary Fig.

7C.

Comment 13: Knockout mice should be placed on HFD or keto diet to test for the effects of PTPMT1 depletion.

We appreciate this thoughtful suggestion. We will certainly incorporate this suggestion into our future studies, expanding beyond the scope of the current initial report.

Comment 14: Suggestions on Fig 4A.

Please see our response to Comment 10.

Comment 15: Suggestions for improving echocardiographs.

These analyses were conducted by trained personnel at the Emory Animal Physiology Core. The data presented in our manuscript was provided by them. We appreciate bringing the issues to our attention, and we will inform them accordingly.

Comment 16: Comment on Fig 5B.

The tissues were sectioned at comparable, if not identical, levels. WT and PTPMT1 knockout heart sections look dramatically different because of the dilated myopathy observed in the knockout hearts.

Comment 17: Comment on Fig 5C.

We believe the cell death occurred predominantly in cardiomyocytes.

Version 2

Abstract

Abstract

While mitochondria in different tissues have distinct preferences for energy sources, they are flexible in utilizing competing substrates for metabolism according to physiological and nutritional circumstances. However, the regulatory mechanisms and significance of metabolic flexibility are not completely understood. Here we report that the deletion of PTPMT1, a mitochondria-based phosphatase, critically alters mitochondrial fuel selection – the utilization of pyruvate, a key mitochondrial substrate derived from glucose (the major simple carbohydrate), is inhibited, whereas the fatty acid utilization is enhanced. PTPMT1 knockout does not impact the development of the skeletal muscle or heart. However, the metabolic inflexibility ultimately leads to muscular atrophy, heart failure, and sudden death. Mechanistic analyses reveal that the prolonged substrate shift from carbohydrates to lipids causes oxidative stress and mitochondrial destruction, which in turn results in marked accumulation of lipids and profound damage in the knockout muscle cells and cardiomyocytes. Interestingly, PTPMT1 deletion from the liver or adipose tissue does not generate any local or systemic defects. These findings suggest that PTPMT1 plays an important role in maintaining mitochondrial flexibility and that their balanced utilization of carbohydrates and lipids is essential for both the skeletal muscle and the heart despite the two tissues having different preferred energy sources.

Introduction

Mitochondria, the powerhouse of the cell, produce energy in the form of ATP from the breakdown of carbohydrates, fats, and proteins. Mitochondrial abundance and their preferences for metabolic substrates differ in various cell types 1. While mitochondria are abundant in red skeletal muscles such as Soleus, and the heart, the mitochondrial content in white muscles, such as Extensor Digitorum Longus (EDL), is much lower. Mitochondria in the skeletal muscle and heart have distinct preferences for energy sources - skeletal muscle mitochondria prefer glucose (the major simple carbohydrate) as the substrate, whereas cardiac mitochondria mainly use fatty acids as the fuel. Moreover, different types of skeletal muscle cells (fibers) utilize glucose to produce energy in different mechanisms - in slow-twitch oxidative muscle fibers glucose (via pyruvate, a critical metabolite derived from glucose) is predominantly oxidized within the mitochondria to generate energy, whereas in fast-twitch muscle fibers pyruvate is primarily reduced to lactate in the cytosol promoting glycolytic flux to rapidly produce energy and support quick contractions. Nevertheless, both muscle and heart mitochondria are flexible in selecting substrates in response to stress (exercise, fasting, etc.) and under disease conditions (heart failure, diabetes, etc.) 2–5. It remains to be determined how mitochondrial utilization of various competing metabolic substrates is coordinated and whether disruption of mitochondrial flexibility in fuel selection affects the function of the heart and skeletal muscle.

PTPMT1, encoded by nuclear DNA, is localized to the mitochondrion and anchored at the inner membrane 6. It dephosphorylates phosphatidylinositol phosphates (PIPs). PIPs are a class of membrane phospholipids that bind to a distinctive set of effector proteins, thereby regulating a characteristic suite of cellular processes, including membrane trafficking and ion channel/transporter functions 7, 8. PTPMT1 is also involved in the synthesis of cardiolipin by converting phosphotidylglecerol phosphate to phosphotidylglecerol, the precursor of cardiolipin 9. Global knockout of PTPMT1 results in developmental arrest and post-implantation lethality 9, 10. Our previous studies suggest that PTPMT1 facilitates mitochondrial metabolism largely by dephosphorylation of downstream PIP substrates 11–13 that appear to inhibit mitochondrial oxidative phosphorylation but enhance cytosolic glycolysis by activation of mitochondrial uncoupling protein 2 (UCP2) 11, a transporter of four-carbon (C4) dicarboxylate intermediates of the tricarboxylic acid (TCA) cycle that has been shown to regulate cellular energetics by limiting mitochondrial oxidation of glucose 14–18. PTPMT1 is highly expressed in the heart and skeletal muscle 10. In the present study, we exploit PTPMT1 tissue-specific knockout models to address the role and mechanisms of PTPMT1-facilitated mitochondrial metabolism in these tissues. We find that PTPMT1 plays an important role in facilitating mitochondrial utilization of carbohydrates and that balanced fuel selection is essential for maintaining both muscle and heart functions despite the two tissues having distinct preferences for energy sources.

Results

Knockout of PTPMT1 from skeletal muscles results in defective contractile function and progressive muscle atrophy

To determine the role of PTPMT1-mediated metabolism in the skeletal muscle and heart, we generated tissue-specific PTPMT1 knockout mice (PTPMT1fl/fl/CKMM-Cre+) by crossing PTPMT1 floxed mice 11 and muscle creatine kinase promoter-driven Cre transgenic mice (CKMM-Cre), which express the Cre recombinase in the skeletal muscle and heart starting at embryonic day 13 19. Quantitative reverse transcription PCR (qRT-PCR) showed ∼95% and ∼80% deletion of PTPMT1 in the skeletal muscle and heart in PTPMT1fl/fl/CKMM-Cre+ mice, respectively (Figure 1A). These animals were indistinguishable from their wild-type (WT, PTPMT1+/+/CKMM-Cre+) littermates up to 3 months of age, suggesting that PTPMT1 deletion did not affect muscle or heart development. The knockout mice exhibited reduced weight gain starting at 4 to 5 months and their body sizes were significantly smaller than those of control mice at 8 months (Figure 1A). Muscle wasting was observed in PTPMT1fl/fl/CKMM-Cre+mice at 8 months or older (Figure 1A). Histopathological examination of skeletal muscles of these knockout mice revealed characteristic changes of muscle atrophy - marked variations in muscle fiber diameters and the presence of ghost fibers or more interstitial space (Figure 1B). In addition, Masson’s trichrome staining revealed increased collagen-rich fibrotic regions in PTPMT1 knockout muscles (Figure 1B). We then examined PTPMT1 knockout mice at 6-8 months (prior to the onset of muscle wasting).

Deletion of PTPMT1 from skeletal muscles results in defective contractility and progressive muscle atrophy.

(A) Representative 8-month-old of PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+mice, and hind limbs and hearts dissected from these mice were photographed. EDL and Soleus (n=8 mice/genotype), and heart (n=4 mice/genotype) weights were measured. PTPMT1 mRNA levels in EDL, Soleus, and heart tissues were determined by quantitative reverse transcription PCR (qRT-PCR) (n=4 mice/genotype). (B) Skeletal muscle sections prepared from 8-month-old PTPMT1+/+/CKMM-Cre+and PTPMT1fl/fl/CKMM-Cre+ mice were processed for H&E staining and Masson’s Trichrome staining. One representative image from 3 mice/genotype is shown. (C) Six-month-old PTPMT1+/+/CKMM-Cre+and PTPMT1fl/fl/CKMM-Cre+ mice (n=7/genotype) were subjected to wire hang tests. Relative forelimb muscle strength was determined. (D and E) Seven to eight-month-old PTPMT1+/+/CKMM-Cre+(n=5 males and 5 females) and PTPMT1fl/fl/CKMM-Cre+(n=3 males and 5 females) mice were assessed by treadmill exercise tests as described in Materials and Methods. Maximum speed (D) and duration of the run (E) were recorded. (F and G) Soleus (F) and EDL (G) dissected from 7 to 8-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM- Cre+ mice (n=3 mice, 6 muscles/genotype) were subjected to ex vivo isometric force measurements. Specific contractile forces produced at the indicated frequencies of stimulation were normalized to the physiological cross-sectional area. Shown on the left are representative absolute forces produced at 1, 10, and 100 Hz. * p < 0.05.

PTPMT1fl/fl/CKMM-Cre+ mice became fatigued faster and fell more quickly than PTPMT1+/+/CKMM- Cre+ control animals in wire hang tests, indicating muscle weakness (Figure 1C). In treadmill exercise tests, both male and female knockout mice displayed reduced maximum speed (Figure 1D), endurance (Figure 1E), and distance of the run (Supplemental Figure 1A). To determine whether the reduced muscle strengths of the knockout mice resulted from muscle cell-intrinsic defects, isometric contractile properties of isolated muscles were examined. Ex vivo force-frequency measurements were performed on Soleus, a more oxidative muscle with a higher percentage of slow-twitch Type I fibers, and EDL, a more glycolytic muscle with a higher percentage of fast-twitch Type II fibers. Both knockout Soleus and EDL showed decreased optimal length, the length of the muscle at which maximal force was achieved (Supplemental Figure 1, C and D). Specific force production of PTPMT1 knockout Soleus (Supplemental Figure 1E) and EDL (Supplemental Figure 1E) was decreased. However, normalized force production was decreased only in knockout Soleus (Figure 1F), but not in knockout EDL (Figure 1G). These data suggest that PTPMT1 plays a more important role in oxidative than glycolytic muscle fibers although both knockout Soleus and EDL developed muscle atrophy subsequently.

To determine the impact of PTPMT1 depletion from the skeletal muscle and heart on the metabolism of the whole body, we measured plasma glucose and fatty acid levels and found that they were comparable in PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice (Supplemental Figure 2A). Plasma levels of lactate (Supplemental Figure 2B) and lipid metabolic products (Supplemental Figure 2C) in PTPMT1 knockout mice were also relatively normal (except that cholesterol and nonesterified fatty acids levels were decreased and marginally increased, respectively, in female knockout mice). Adipokine array assays of the plasma showed no or subtle differences in Adiponectin, Leptin, Lipocalin-2, ANGPTL3, Resistin, FGF-21, DPPIV, Fetuin A, IGF-1, etc. between PTPMT1 knockout and control mice (Supplemental Figure 2D). In addition, no differences in glucose tolerance tests were observed between PTPMT1 knockout and control mice (Supplemental Figure 2E). Furthermore, we assessed the expression levels of key enzymes involved in glucose and lipid metabolism in PTPMT1 knockout muscles. No changes in HK2, PKM1, LDHA, DICER1, MCAD (ACADM), ACADL, HADHA, CPT2, or FABP3 were found. However, the expression of CPT1B, was increased in PTPMT1 knockout muscles (Supplemental Figure 2, F and G). These results suggest that systemic glucose and lipid metabolism in these tissue-specific knockout mice was not significantly changed, but fatty acid partitioning for oxidation in the mitochondria was elevated in PTPMT1 knockout muscles.

PTPMT1 depletion ultimately leads to mitochondrial damage and bioenergetic stress in knockout skeletal muscles

Gömöri trichrome staining revealed ragged red fibers in PTPMT1 knockout skeletal muscles (Figure 2A), as observed in various types of human mitochondrial myopathies. Electron microscopic analyses showed that interfibrillar mitochondria in knockout Soleus and EDL were disorganized and enlarged, in contrast to normal mitochondria that typically reside in pairs and position on either side of the Z-disc in control muscles. In addition, there was a massive accumulation of subsarcolemmal mitochondria in the knockout muscles, whereas the content of intermyofibrillar mitochondria decreased (Figure 2B). Although EDL, unlike Soleus, uses more cytosolic glycolysis than mitochondrial oxidative phosphorylation for energy production, the mitochondrial structural changes appeared to be more severe in EDL than Soleus in the knockout mice (Figure 2B). Furthermore, a massive accumulation of intramyofibrillar lipids were detected by Oil Red O staining in 6-month or older PTPMT1 knockout muscles (Figure 2C).

PTPMT1 loss ultimately leads to abnormal mitochondrial distribution, structural damage, and bioenergetic stress in skeletal muscles.

(A) Skeletal muscle sections prepared from 6-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM-Cre+ mice were processed for Gömöri trichrome staining. One representative image from 3 mice/genotype is shown. (B) Soleus and EDL dissected from 8-month-old PTPMT1fl/fl/CKMM-Cre+and PTPMT1+/+/CKMM- Cre+ mice were processed for transmission electron microscopic examination. One representative image from 3 mice/genotype is shown. (C) Skeletal muscle sections prepared from 6-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM-Cre+ mice were processed for Oil Red O staining to visualize lipids. One representative picture from 3 mice/genotype is shown. (D) Total DNA was extracted from Soleus and EDL dissected from 8-month-old PTPMT1fl/fl/CKMM-Cre+and PTPMT1+/+/CKMM-Cre+ mice (n=3/genotype). Mitochondrial content was estimated by comparing the mitochondrial gene cytochrome B DNA levels to the nuclear gene 18S DNA levels by qPCR. (E) Total ATP levels in Soleus and EDL dissected from 8-month-old PTPMT1fl/fl/CKMM-Cre+and PTPMT1+/+/CKMM-Cre+ mice (n=6/genotype) were determined. (F) Whole cell lysates prepared from the Soleus and EDL isolated from 7-month- old PTPMT1fl/fl/CKMM-Cre+ and their control mice were examined by immunoblotting with the indicated antibodies. Representative results from 3 mice/genotype are shown.

Total mitochondrial content in PTPMT1 knockout Soleus and EDL also slightly decreased compared to that in control muscles (Figure 2D). ATP levels in both knockout Soleus and EDL decreased significantly compared to WT counterparts (Figure 2E). Consistent with these data, the energetic stress sensor, AMP- activated kinase (AMPK), was highly activated in PTPMT1fl/fl/CKMM-Cre+ muscles as determined by the phosphorylation levels of Thr172 (Figure 2F). Acetyl-CoA carboxylase (ACC), a negative regulator of fatty acid oxidation, was inhibited, as evidenced by a marked increase in the inhibitory phosphorylation of this enzyme (Figure 2F), implying increased fatty acid oxidation in PTPMT1-depleted mitochondria. In agreement with previous findings that AMPK negatively regulates mTOR signaling 20, 21, mTOR was substantially inhibited. Activities of S6K, S6, and 4E-BP1, key downstream components of mTOR signaling, were also concomitantly decreased (Figure 2F), indicating diminished anabolic activities and consistent with the muscle atrophy phenotype. Interestingly, Akt activities, as reflected by its phosphorylation levels (Ser473 and Thr308), were increased in PTPMT1 knockout skeletal muscles, recapitulating mTOR and raptor (a component of the mTOR complex 1) knockout muscles 22, 23. Likely due to the cross-talk between Akt and Ras signaling pathways, Erk activities were also slightly increased in these knockout muscles (Figure 2F).

Given that muscle atrophy developed in PTPMT1fl/fl/CKMM-Cre+mice after 6-8 months, to assess the direct impact of PTPMT1 deletion on mitochondrial metabolism, it is important to examine mitochondria in young PTPMT1 knockout mice before the tissue damage. Expression levels of electron transport chain complexes in these PTPMT1-ablated mitochondria were not changed (Supplemental Figure 3A). Total cellular ATP levels in PTPMT1 knockout muscles at this age were comparable to those in control tissues (Supplemental Figure 3B), and the knockout muscles did not show bioenergetic/metabolic stress (Supplemental Figure 3C). We also checked LC3-I/II levels in PTPMT1 knockout muscles and found no evidence of elevated autophagic activities (Supplemental Figure 3D). These results indicate that the decrease in total cellular ATP levels in older knockout muscles (Figure 2E) was a secondary not direct effect of PTPMT1 depletion.

PTPMT1 loss impairs mitochondrial utilization of pyruvate, whereas the fatty acid utilization is enhanced

We next sought to address the mechanisms by which PTPMT1 loss impacts mitochondrial metabolism. To avoid secondary effects, we examined mitochondrial metabolic activities of PTPMT1 knockout muscles in 3-month-old PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice. Knockout muscle tissues showed decreased maximal reserve oxidative capacities in the presence of full substrates (Figure 3A), consistent with our previous observations from other PTPMT1 knockout cell types 11, 12, 24. In addition, we isolated mitochondria from young skeletal muscles and assessed mitochondrial metabolism in the presence of ADP and a single metabolic substrate by real-time measurement of ATP synthesis- driven oxygen consumption. When pyruvate, a critical metabolite derived from glucose, was provided as the sole substrate, ATP synthesis-driven oxygen consumption in PTPMT1-deficient mitochondria was markedly decreased (Figure 3B), suggesting that mitochondrial utilization of pyruvate was diminished in the absence of PTPMT1. Interestingly, we observed enhanced oxygen consumption in these mitochondria when fatty acids were supplied as the fuel (Figure 3C). A slightly increased respiratory response in PTPMT1-depleted mitochondria was also detected when glutamate, another mitochondrial substrate, was provided (Figure 3D). Importantly, feeding with succinate, the substrate for Complex II of the electron transport chain, resulted in similar oxygen consumption in PTPMT1 null and control mitochondria (Figure 3E), validating the intact function of the electron transport chain in the knockout mitochondria and excluding flavin adenine dinucleotide and quinone-linked deficiencies. These mechanistic data suggest that PTPMT1 plays an important role in facilitating carbohydrate oxidation in mitochondria.

PTPMT1 ablation impairs mitochondrial utilization of pyruvate, whereas the fatty acid utilization is enhanced.

(A) Muscle cross-sections prepared with biopsy punches from PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice (n=4/genotype) at 4 months of age were measured for OCRs at the basal level and following the addition of oligomycin (8 μM), FCCP (4 μM), and antimycin A/rotenone (1 μM). (B-E) Mitochondria were isolated from the skeletal muscles dissected from PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice (n=3/genotype) at 3 months of age. Mitochondrial oxygen consumption (10 μg of mitochondrial protein) was measured in the presence of pyruvate (5 mM)/malate (5 mM) (B), palmitoyl-CoA (40 μM)/carnitine (40 μM)/malate (5 mM) (C), glutamate (5 mM)/malate (5 mM) (D), or succinate (10 mM) (E), following the addition of ADP (4 mM), oligomycin (1.5 μM), FCCP (4 μM), and antimycin A/rotenone (1 μM). Experiments were repeated three times with three independent pairs of mice. Similar results were obtained in each experiment. (F-H) Levels of pyruvate (F), α- KG (G), and acetyl-CoA (H) in the lysates of the mitochondria isolated from the above skeletal muscles were measured (n=6/genotype). (I) Mitochondria freshly isolated from the skeletal muscles of PTPMT1fl/fl/CKMM-Cre+ and PTPMT1+/+/CKMM-Cre+ mice (n=4/genotype) were washed three times in MAS buffer. The mitochondria were then incubated with pyruvate (5 mM)/malate (5 mM) and ADP (4 mM) at 37°C. Five min later, mitochondria were collected, washed, and lysed. α-KG levels in the mitochondrial lysates were measured. (J) Pyruvate dehydrogenase (PDH) activities in the mitochondrial lysates were determined (n=5-6 mice/genotype). (K) MYH4, MYH2, and MYH7 mRNA levels in the skeletal muscles dissected from 3-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM-Cre+mice (n=4/genotype) were determined by qRT-PCR.

Consistent with the above notion, intramitochondrial pyruvate levels in the mitochondria freshly isolated from PTPMT1 knockout muscles decreased by half (Figure 3F). Levels of α-ketoglutarate (α-KG), an important metabolite in the TCA cycle, also decreased in the knockout mitochondria (Figure 3G), although steady-state mitochondrial acetyl-CoA levels were not changed (Figure 3H). To further determine if pyruvate transport efficiency may be reduced in PTPMT1-ablated mitochondria, we incubated fresh mitochondria with pyruvate and measured acute production of α-KG from extramitochondrial pyruvate. As shown in Figure 3I, α-KG production in PTPMT1-ablated mitochondria was indeed decreased compared to that in control mitochondria. Notably, the activity of pyruvate dehydrogenase (PDH), the enzyme that converts pyruvate to acetyl-CoA to feed the TCA cycle, was not affected in the knockout mitochondria (Figure 3J). These observations together with that the mitochondrial pyruvate carrier/transporter (MPC) was comparably expressed in PTPMT1-ablated mitochondria (Figure 2F) suggest that PTPMT1 null mitochondria were impaired in uptaking pyruvate. Possibly due to an adaptive response to the inhibited pyruvate oxidation in the mitochondria, skeletal muscle fiber-type switching occurred in PTPMT1fl/fl/CKMM-Cre+ mice. Muscle fibers in these knockout mice switched from oxidative to glycolytic fibers, as evidenced by a dramatic increase (∼170-fold) in the levels of MYH4, a marker of glycolytic fast-twitch Type 2B fibers in PTPMT1 knockout mice (Figure 3K).

PTPMT1fl/fl/CKMM-Cre+ mice manifest late onset cardiac dysfunction

The heart in PTPMT1fl/fl/CKMM-Cre+ mice was less impacted than the skeletal muscle although PTPMT1 was ∼80% deleted from the heart (Figure 1A). At 7 months, when skeletal muscles in these mice displayed atrophy and myocyte damage, no evident histological changes were observed in heart tissues (Supplemental Figure 4A). Echocardiographic examination showed no difference in left ventricle (LV) contractility in PTPMT1 knockout mice as reflected by similar LV fractional shortening (FS), Ejection Fraction (EF), and the ratio of peak velocity of early to late filling of mitral inflow (E/A) between knockouts and control mice (Supplemental Figure 4B). Mitochondrial content in PTPMT1fl/fl/CKMM- Cre+heart tissues was comparable to that in PTPMT1+/+/CKMM-Cre+control tissues (Supplemental Figure 4C). Total ATP levels in PTPMT1 knockout heart tissues were marginally but not significantly decreased (Supplemental Figure 4D). Moreover, mitochondria in knockout cardiomyocytes showed minimal structural changes (Supplemental Figure 4E). Consistent with these observations, no bioenergetic/metabolic stress was detected in the knockout cardiomyocytes according to the activation status of AMPK and mTOR signaling (Supplemental Figure 4F).

However, severe cardiac dysfunction was observed in 10 to 12-month-old PTPMT1fl/fl/CKMM-Cre+ mice. Histopathological examination revealed that cardiomyocytes in these animals had uneven sizes and were disorganized (Figure 4A). In addition, fibrotic lesions, indicative of myocardial damage, were observed. Echocardiographic examination showed that LV contractility, as measured by FS and EF in M-mode echocardiographic tracing of LV (Figure 4, B-D) and recorded by M-Mode echocardiography movies (Supplemental Video 1, A and B), was markedly decreased in PTPMT1 knockout mice. The E/A ratio determined in pulsed-wave Doppler recording of mitral valve inflow was doubled in the PTPMT1fl/fl/CKMM-Cre+ heart (Figure 4, E and F). Collectively, these data suggest that PTPMT1 depletion also impaired cardiac function albeit later than skeletal muscle dysfunction.

PTPMT1fl/fl/CKMM-Cre+mice manifest late-onset cardiac dysfunction.

(A) Heart tissue sections prepared from 12-month-old PTPMT1+/+/CKMM-Cre+and PTPMT1fl/fl/CKMM- Cre+ mice were processed for H&E staining. One representative image from 3 mice/genotype is shown. (B-F) Cardiac morphology and function of 12-month-old PTPMT1+/+/CKMM-Cre+ and PTPMT1fl/fl/CKMM-Cre+ mice (n=5/genotype) were examined by echocardiography. Representative long-axis views in M-mode echocardiographic tracing of LV are shown (B). LV FS (C) and LV ejection fraction (EF) (D) were determined. Representative pulsed-wave Doppler recordings of mitral valve inflow are shown (E). Ratios of peak velocity of early to late filling of mitral inflow (E/A) were determined (F).

Heart-specific deletion of PTPMT1 leads to dilated cardiomyopathy and heart failure

Given that PTPMT1 was only ∼80% deleted from the heart in PTPMT1fl/fl/CKMM-Cre+ mice and that impaired skeletal muscle function in these animals may potentially complicate the heart phenotypes, to further investigate the role of PTPMT1-mediated metabolism specifically in cardiac muscles, heart- specific PTPMT1 knockout mice (PTPMT1fl/fl/MYH6-Cre+) were generated by crossing PTPMT1 conditional mice 11 and MYH6-Cre transgenic mice, which express the Cre recombinase specifically in cardiomyocytes 25. These knockout mice appeared healthy in the first several months although PTPMT1 was nearly completely deleted from the heart (but not in Soleus and EDL) (Supplemental Figure 5A). No obvious tissue morphological changes or fibrotic lesions were found in PTPMT1 knockout hearts (Supplemental Figure 5B), confirming that PTPMT1-mediated metabolism is dispensable for the development of the heart. Echocardiography also showed normal heart function in 3-month-old PTPMT1fl/fl/MYH6-Cre+ mice, as evidenced by normal FS, EF, and E/A ratios (Supplemental Figure 5, C- E). Knockout mice did not exhibit any defects during treadmill exercises at 6 months of age (Supplemental Figure 5F). However, all PTPMT1fl/fl/MYH6-Cre+ knockout mice died at 10-16 months, and they often succumbed suddenly (Figure 5A) although their body weights were relatively normal. PTPMT1 knockout hearts showed enlarged ventricular chambers. Severe structural damage in cardiac myocytes and tremendous fibrotic lesions were observed in the knockout hearts (Figure 5B). Immunostaining of cleaved caspase 3 illustrated increased apoptosis in PTPMT1fl/fl/MYH6-Cre+cardiomyocytes as compared to control cells (Figure 5C). Echocardiography revealed a profound dilated cardiomyopathy and heart failure in PTPMT1fl/fl/MYH6-Cre+mice, including both systolic and diastolic LV dilation, thinning of ventricular walls, depression of EF and FS, arrhythmias, and impaired myocardial contraction (Figure 5, D-J and Supplemental Video 2, A and B). Cardiac strain images demonstrated that the global function of PTPMT1fl/fl/MYH6-Cre+hearts, as reflected by global longitudinal strain, was significantly decreased (Supplemental Figure 6, A and B), similar to that observed in PTPMT1fl/fl/CKMM-Cre+mice.

Deletion of PTPMT1 from the heart ultimately leads to dilated cardiomyopathy and heart failure.

(A) Kaplan-Meier survival curves of PTPMT1fl/fl/MYH6-Cre+(n=18) and PTPMT1+/+/MYH6-Cre+ mice (n=20). (B) Heart tissue sections prepared from 10 to 12-month- old PTPMT1fl/fl/MYH6-Cre+and PTPMT1+/+/MYH6-Cre+ mice (n=4/genotype) were processed for H&E staining and Masson’s Trichrome staining. Representative images are shown. (C) Heart tissue sections prepared from 11-month-old PTPMT1fl/fl/MYH6-Cre+and PTPMT1+/+/MYH6-Cre+ mice were processed for immunofluorescence staining for cleaved caspase 3 followed by DAPI counterstaining. One representative image from 3 mice/genotype is shown. (D-J) Eleven to twelve-month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice (n=5/genotype) were examined by echocardiographic evaluation of ventricular function. Representative long- (upper panel) and short- (lower panel) axis views in M-mode echocardiographic tracing of LV (D), and pulsed-wave Doppler recordings of mitral valve inflow (E) are shown. Left ventricular internal diameters at diastole (LVID-d) and systole (LVID-s) were measured (F). LV ejection fraction (EF) and fractional shortening (FS) (G), ratios of peak velocity of early to late filling of mitral inflow (E/A) (H), the thickness of LV anterior wall at the end of diastole (LVAW-d) and the end of systole (LVAW-s) (I), and thickness of LV posterior wall at the end of diastole (LVPW-d) and at the end of systole (LVPW-s) (J) were determined.

Unlike those in younger knockout mice, total cellular ATP levels in 10 to 12-month-old PTPMT1fl/fl/MYH6-Cre+ hearts were decreased by half (Figure 6A), and the knockout cardiomyocytes exhibited pronounced energetic stress as demonstrated by a substantial increase in p-AMPK (T172) and decrease in p-mTOR (S2448) (Figure 6B). To determine the direct effects of PTPMT1 loss on cardiomyocyte metabolism, we examined heart tissues isolated from young (3 months old) knockout mice before cardiomyocyte and mitochondria damages occurred. The mitochondrial content in these young PTPMT1 knockout cardiac tissues was comparable to that in control tissues (Supplemental Figure 7A), and there were no changes in the expression levels of mitochondrial complexes (Supplemental Figure 7B). Moreover, no bioenergetic/metabolic stress was detected in the knockout cardiomyocytes at this age (Supplemental Figure 7C). However, like PTPMT1 knockout skeletal muscle mitochondria (Figure 3, B-D), PTPMT1-ablated cardiac mitochondria also showed altered fuel selection – the utilization of pyruvate was decreased (Figure 6C), whereas the free fatty acid and glutamate utilization was increased in PTPMT1fl/fl/MYH6-Cre+cardiac mitochondria (Figure 6, D and E). Intramitochondrial pyruvate levels in PTPMT1-ablated cardiac mitochondria decreased (Figure 6F) while acetyl-CoA levels were not changed (Supplemental Figure 7D), consistent with the metabolite data obtained from PTPMT1- depleted skeletal muscle mitochondria (Figure 3, F and H). Furthermore, acute production of α-KG from extramitochondrial pyruvate was decreased in the cardiac mitochondria lacking PTPMT1 (Figure 6G), although PDH activity in these mitochondria was not significantly affected (Figure 6H).

PTPMT1 deficiency causes mitochondrial substrate shift and oxidative stress, leading to mitochondrial damage and lipid accumulation in PTPMT1 knockout cardiomyocytes.

(A) Total ATP levels in the heart tissues dissected from PTPMT1fl/fl/MYH6- Cre+and PTPMT1+/+/MYH6-Cre+ mice at the indicated ages (n=5 mice/genotype). (B) Whole cell lysates prepared from the heart tissues of 10 to 12-month-old PTPMT1fl/fl/MYH6-Cre+and PTPMT1+/+/MYH6-Cre+ mice were examined by immunoblotting with the indicated antibodies. Representative results from 3 mice/genotype are shown. (C-E) Oxygen consumption of the mitochondria isolated from the heart tissues of 2 to 3-month-old PTPMT1fl/fl/MYH6-Cre+and PTPMT1+/+/MYH6-Cre+ mice (n=3/genotype) was measured in the presence of pyruvate (5 mM)/malate (5 mM) (C), palmitoyl-CoA (40 μM)/carnitine (40 μM)/malate (5 mM) (D), or glutamate (5 mM)/malate (5 mM) (E), following the addition of ADP (4 mM), oligomycin (1.5 μM), FCCP (4 μM), and antimycin A/rotenone (1 μM). OCRs at basal levels, in response to ADP addition, and maximal reserve capabilities were determined. (F) Levels of pyruvate in the lysates of the cardiac mitochondria isolated from 2 to 3-month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice (n=6/genotype) were determined. (G and H) Mitochondria isolated above were washed three times in MAS buffer and then incubated with pyruvate (5 mM)/malate (5 mM) and ADP (4 mM) at 37°C. Five min later, the mitochondria were collected, washed, and lysed. α-KG levels in the mitochondrial lysates were measured (n=4 mice/genotype) (G). Pyruvate dehydrogenase (PDH) activities in the mitochondrial lysates were determined (n=5 mice/genotype) (H). (I) Heart tissue sections prepared from 3-month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+mice (4 mice/genotype) were processed for ROS staining. One representative picture from 4 mice/genotype is shown. (J) Heart tissues dissected from 6 to 8- month-old PTPMT1fl/fl/MYH6-Cre+ and PTPMT1+/+/MYH6-Cre+ mice were processed for TEM examination. One representative image from 4 mice/genotype is shown. (K) Heart tissue sections prepared from 10 to 12-month-old PTPMT1fl/fl/MYH6-Cre+and PTPMT1+/+/MYH6-Cre+ mice were processed for Oil Red O staining to visualize lipids. One representative picture from 3 mice/genotype is shown.

Oxidation of fatty acids in mitochondria yields much more ATP relative to carbohydrates (via pyruvate); however, fatty acid oxidation requires a greater rate of oxygen consumption for a given rate of ATP synthesis than carbohydrates. Such that the byproduct reactive oxygen species are increased when mitochondria switch to fatty acids for energy metabolism. Indeed, PTPMT1 knockout heart tissues showed elevated overall cellular ROS levels even at 3 months (Figure 6I). Persistent oxidative stress can cause mitochondrial damage. Electron microscopic examination revealed that the myocardium of PTPMT1fl/fl/MYH6-Cre+mice was distinguishable from that of PTPMT1+/+/MYH6-Cre+control mice by the structures of mitochondria at 6-8 months (Figure 6J). Similar to that observed in PTPMT1 knockout skeletal muscle cells (Figure 2C), drastic accumulation of lipids or lipid intermediates was observed in 10- 12-month-old PTPMT1fl/fl/MYH6-Cre+ cardiomyocytes (Figure 6K), which is likely responsible for the increased apoptosis and fibrillation in the knockout heart (Figure 5, B and C) due to the toxicity of excess lipids (lipotoxicity) 26, 27.

PTPMT1 is dispensable for the liver and adipose tissues

Finally, to further determine whether PTPMT1-facilitated carbohydrate oxidation in mitochondria is also important for other important metabolic tissues, we generated liver-specific (PTPMT1fl/fl/ALB-Cre+) and fat-specific (PTPMT1fl/fl/Adipoq-Cre+) PTPMT1 knockout mice by crossing PTPMT1 floxed mice 11 with ALB-Cre 28 and Adipoq-Cre 29 mice, which express Cre specifically in hepatocytes and adipocytes, respectively. In contrast to muscle/heart- and heart-specific knockout mice, liver-specific PTPMT1 knockout mice were healthy without overt abnormalities. They had a normal lifespan. Their body weights, liver weights, and blood glucose levels under both feeding and fasting conditions were comparable to those in control animals (Supplemental Figure 8, A-D). Moreover, no significant differences were observed in glucose tolerance tests between PTPMT1fl/fl/ALB-Cre+and control mice (Supplemental Figure 8E). Unlike muscle/heart- and heart-specific PTPMT1 knockout mice in which prominent lipid accumulation was detected in muscle and cardiac myocytes, liver-specific PTPMT1 knockout mice did not display this secondary effect in hepatocytes, and there was no evident tissue damage in the liver or a fatty liver phenotype (Supplemental Figure 8F). Similarly, adipocyte-specific PTPMT1 knockout mice were also healthy. Their body weights and glucose tolerance were comparable to those of control animals (Supplemental Figure 9, A and B). No histological changes were observed in PTPMT1 knockout white adipose and brown adipose tissues (Supplemental Figure 9C). The fact that PTPMT1fl/fl/ALB-Cre+and PTPMT1fl/fl/Adipoq-Cre+ mice well tolerated PTPMT1 deficiency in hepatocytes or adipocytes suggests that the detrimental effects of PTPMT1 depletion-induced mitochondrial substrate shift are highly cell type-specific.

Discussion

Mitochondrial flexibility in substrate selection is essential for maintaining cellular energy homeostasis in physiology and under stress; however, the regulatory mechanisms and its significance for various tissues and cell types remain poorly characterized 1. In this study, we have identified an important role of the mitochondrial phosphatase PTPMT1 in maintaining mitochondrial flexibility. The loss of PTPMT1 decreased mitochondrial utilization of pyruvate, a key mitochondrial substrate derived from glucose, whereas fatty acid oxidation was enhanced (Figure 3C, Figure 6D). Importantly, this prolonged mitochondrial substrate shift causes oxidative stress and mitochondrial damage, ultimately leading to the accumulation of lipids/lipid intermediates in PTPMT1 knockout skeletal muscle and cardiac myocytes. Excess lipids/lipid intermediates are known to be toxic to cardiomyocytes (lipotoxicity) 26, 27. The profound damages in PTPMT1 knockout hearts and muscles that occurred subsequently were likely attributed to these accumulated lipids, although other factors, such as decreased anabolic activities (as demonstrated by diminished mTOR signaling) (Figure 2F, Figure 6B), could also contribute to these phenotypes. The mitochondrial and tissue damages that subsequently occurred, combined with the resulting energetic stress, finally led to muscle atrophy and heart failure in the knockout mice.

One of the important findings in this report is that persistent mitochondrial fuel switch from pyruvate to fatty acids is detrimental for both cardiac and skeletal muscles despite the two tissues having different preferred energy sources. Glucose is the major energy source for skeletal muscles. It is used for energy production mainly in mitochondria in oxidative red muscles or through cytosolic glycolysis in glycolytic white muscles. Glucose is also the main energy source for the developing heart, but the adult heart primarily utilizes fatty acids to produce ATP, although the adult heart can also utilize glucose and lactate, the end product of glycolysis, as energy sources. Indeed, when mitochondrial utilization of pyruvate was inhibited by PTPMT1 depletion, heart dysfunction developed much later than skeletal muscle dysfunction. Interestingly, our data from these animal models suggest that efficient carbohydrate oxidation in mitochondria and balanced mitochondrial fuel selection are critically important for both oxidative and glycolytic muscles as well as the adult heart, but not for the development of the skeletal muscle and heart although PTPMT1 depletion took place during the embryonic stage. Undoubtedly, the late-onset phenotypes observed in the knockout mice over time was attributed to the metabolic defects arising from the loss of PTPMT1 in the embryos. Although PTPMT1 knockout muscle cells and cardiomyocytes initially maintained energy homeostasis through enhanced fatty acid and glutamate oxidation, along with metabolic adaptations or activation of alternative energy-producing pathways in the first few months, they eventually encountered substantial energy deficits. This was attributed to the subsequent occurrence of oxidative stress and mitochondrial damage.

PTPMT1 knockout hepatocytes or adipocytes did not display such secondary effects (mitochondria and cell damages). This is possibly because the requirement of carbohydrate oxidation for energy production in hepatocytes and adipocytes is lower than that in skeletal muscle cells and cardiomyocytes, and/or PTPMT1 knockout hepatocytes and adipocytes did not need to uptake excessive fatty acids in the effort to maintain energy homeostasis. Another possibility is that the capabilities of hepatocytes and adipocytes to process fatty acids and to clear lipid metabolites (through cholesterol synthesis and cholesterol exportation out of the cell) are much stronger than muscle cells and cardiomyocytes 30.

Another important finding in this report is that PTPMT1 plays a major role in governing the metabolic fate of pyruvate and maintaining mitochondrial flexible substrate utilization. Pyruvate is at the central branch point of mitochondrial oxidation and cytosolic glycolysis in glucose metabolism − It can enter the mitochondrion for oxidation or be reduced to lactate in the cytosol. Our data suggest that inhibited mitochondrial utilization of pyruvate is most likely responsible for the remodeled metabolism in PTPMT1 knockout cells since ATP synthesis-driven oxygen consumption of PTPMT1-ablated mitochondria was greatly reduced with pyruvate as the sole substrate (Figure 3B, Figure 6C). Moreover, acute production of α-KG from extramitochondrial pyruvate was indeed decreased (Figure 3I, Figure 6G).

The precise mechanism by which PTPMT1 facilitates mitochondrial utilization/uptake of pyruvate remains to be further determined. PTPMT1 is localized to the inner mitochondrial membrane where pyruvate transporters reside. This phosphatase dephosphorylates phosphatidylinositol phosphates (PIPs)11–13. We have shown previously that PIPs directly promote fatty acid-induced activation of UCP2 11. UCP2, unlike UCP1, is a C4 metabolite transporter 14. Elevated UCP2 activity inhibits mitochondrial oxidation of glucose but enhances cytosolic glycolysis although the underlying mechanisms remain unclear 15–18. PTPMT1 depletion conceivably leads to the buildup of downstream PIP substrates in the mitochondrial inner membrane, which in turn remodel mitochondrial energy source selection by enhancing UCP2 function 11 [UCP2 levels in PTPMT1-ablated mitochondria were not changed (Supplemental Figure 3A, 7B)]. It may also be possible that the accumulated PIPs directly inhibit the mitochondrial pyruvate transporter MPC. It is important to note that PTPMT1 was reported to be involved in the synthesis of cardiolipin 9 which is one of the major components of the mitochondrial inner membrane and that plays an important role in maintaining mitochondrial membrane stability and dynamics. Thus, it may also be possible that PTPMT1 depletion decreases mitochondrial pyruvate utilization through reduced cardiolipin synthesis. Comprehensive metabolomic analyses for PTPMT1- deleted cells would provide additional insights into the mechanisms underlying the metabolic alterations caused by PTPMT1 loss.

In summary, in this report we provide evidence that the mitochondria-based multi-functional phosphatase PTPMT1 plays an important role in facilitating mitochondrial utilization of pyruvate and maintaining mitochondrial flexibility for balanced substrate selection. Prolonged mitochondrial energy sources switch from carbohydrates to lipids such as that caused by PTPMT1 depletion is detrimental for the skeletal muscle and heart but not the liver and adipose, ultimately leading to muscle atrophy and heart failure.

Materials and methods

Mice

PTPMT1fl/+ mice with the C57Bl/6 background were generated in our previous study 11. CKMM-Cre+ 19 (Stock# 006475), MYH6-Cre+ 25 (Stock# 011038), ALB-Cre+ 28 (Stock# 003574), and Adipoq-Cre+29 (Stock# 010803) mice with the C57BL/6 background were purchased from the Jackson Laboratory. To avoid potential side effects of Cre, PTPMT1+/+/Cre+mice were used as controls for PTPMT1fl/fl/Cre+ mice and the Cre transgene was hemizygous in both mouse types. In initial pilot experiments, we examined PTPMT1+/+/Cre+ mice and no defects were observed in these animals. Mice were kept under specific pathogen-free conditions in Case Western Reserve University Animal Resources Center and Emory University Division of Animal Resources. Mice of the same age, sex, and genotype were randomly grouped for subsequent analyses (investigators were not blinded). All animal procedures complied with the NIH Guidelines for the Care and Use of Laboratory Animals and were approved by the Institutional Animal Care and Use Committee.

Antibodies

The sources and catalog numbers of the antibodies utilized in this study are listed below.

Wire hang test

Maximal muscular strength of mouse forelimbs was assessed by wire hang tests following a standard protocol. Mice were allowed to grasp and hang on a wire with forelimbs. Time duration when the mouse was able to suspend before falling was recorded. Three trials were carried out for each mouse. The average time was normalized against body weight. Values of control mice were set to 100%.

Treadmill exercise test

Treadmill exercise tests were conducted following the protocol previously reported by trained personnel at Case Western Reserve University Mouse Metabolic Phenotyping Center Analytical Core 31, 32. Adjustable parameters on the treadmill include belt speed (0-99 m/min) and angle of inclination (0-45”). After acclimation for 1 hour, 2.5 m/min incremental increases in treadmill belt speed and 2’ increments in an angle of inclination were performed every 3 min until the mouse exhibited signs of exhaustion. Exhaustion was defined as the mouse spending >50% of the time or X5 sec consecutively on the shock grid. The maximum speed and duration of the run were recorded at the time of exhaustion.

Ex vivo muscle contractility assay

Intact Soleus and EDL isolated from mice were used for the collection of isometric force data using an eight-chamber system 12, 33. Data were analyzed with ADInstruments PowerLab software customized for these experiments. Muscles were first equilibrated for 20 min to mimic conditions of normal activity (low duty cycle, ∼1%). Muscles were then subjected to the length-force relationship test to determine the optimal length at which maximal force is achieved. Muscles were stimulated with frequencies ranging from 1 to 130 Hz to generate force versus frequency relationships. From these relationships, the frequency producing maximal tetanic force (Tmax) was then used for the rest of the experiments. Muscles were stimulated every minute with Tmax for 5–10 min to assure that the preparation was stable at which point the tissue bathing buffer was replaced with one containing 0 mM Ca2+ + 0.1 mM EGTA. The muscles were then stimulated every minute at Tmax to determine the impact of the removal of external calcium on muscle force. After 20 min, the 0 mM Ca2+ buffer was washed out and replaced by a buffer containing 2.5 mM Ca2+, and the muscles were stimulated every minute for 20 min at Tmax to allow for force recovery. After force measurement protocols, muscle dimensions and masses were measured for the determination of cross-sectional normalized forces (relative force). Muscle force was reported as absolute force (mN) and force normalized to the following muscle physiological cross-sectional area (PCSA, N/cm2) equation as previously reported 34, 35 with a small modification in that in our final length calculation, we considered the length to weight ratio of muscle fibers for stretched muscles (tendon-to-tendon) of ex vivo muscles in contractility chambers 36 as the actual muscle fiber size was smaller than the length of the measured stretched muscle. Since the PCSA is very sensitive to length, by adjusting to the length of the muscle fiber, force estimation was more precise. Thus, our final formula was: Muscle force (N/cm2) = (force (g) x (muscle length (cm) x 0.75 (EDL) or 0.85 (Soleus) x 1.06 (muscle density))/(muscle weight (g) x 0.00981).

Echocardiography

Echocardiographic studies were performed by trained investigators at Emory University Department of Pediatrics Animal Physiology Core as previously reported 37, 38. Mice were anesthetized with 1-2% isoflurane/100% oxygen continuously and positioned on a warming platform set to 37°C. The heart rate was controlled at 360-460 beats per minute to standardize recordings. ECG and respiratory rate were monitored for physiological assessment while scanning. A Vevo 2100 digital high-frequency ultrasound system (FujiFilm Visualsonics Inc, Toronto, ON, Canada) equipped with an MS400, 30-MHz transducer was used to acquire the images. Interventricular septal thicknesses (IVS), left ventricular internal dimensions (LVID) and posterior wall thicknesses (LVPW) at diastole and systole (IVS-d, LVID-d, LVPW-d and IVS-s, LVID-s, LVPW-s) were measured from M-mode images in both parasternal long axis and short-axis views. Left ventricle volume at diastole and systole (LV Vol-d, LV Vol-s), left ventricular ejection fraction (LVEF) and left ventricular fractional shortening (LVFS) were calculated as the following formula (Vevo 2100 Workstation Software VisualSonics INC, Toronto, ON, Canada): LV Vol- d(ul) = [(7.0 /(2.4 + LVID-d)] * LVID-d3); LV Vol-d(ul) = [(7.0 /(2.4 + LVID-s)] * LVID-s3); EF (%) = 100 * [(LV Vol-d – LV Vol-s) / LV Vol-d]; FS (%) = 100 * [(LVID-d – LVID-s) / LVID-d]. The mitral valve flow Doppler velocities were measured by the apical four-chamber view. Early (E) and late (A) mitral valve inflow velocity and mitral valve E to A ratios (MV E/A) were also obtained.

Echocardiographic speckle-tracking based strain measurement was also performed using the VevoStrainTM application (VisualSonics Vevo 2100 Imaging System, FujiFilm VisualSonics, Inc, Toronto, ON, Canada) which produces velocity strain and time-to-peak analyses on myocardial wall images. The B-Mode cine loop was acquired for Vevo Strain analysis. The qualified consecutive cardiac cycles were selected between one respiration cycle and the next in the EKG panel. Both endocardium and epicardium were traced simultaneously and semi-automatically at the parasternal long-axis view. VevoStrain builds the dynamic LV wall trace for all frames in the cine loop. Velocity, displacement, strain, and strain rate data were calculated based on a speckling-tracking algorithm provided by the Vevo software. Data on overall cardiac tissue performance, segmental time-to-peak, and segmental phase quantification were also collected.

Intracellular ATP quantification

Skeletal muscle and heart tissues were lysed in lysis buffer [Tris-HCl (50 mM), pH 7.4, NP-40 (1%), Na- deoxycholate (0.25%), NaCl (150 mM), EDTA (1 mM), NaF (1 mM), Na3VO4 (1 mM), PMSF (1 mM), and Aprotinin/Leupeptin (10 µg/ml)] for 30 min. Samples were then centrifuged (10,000g) for 10 min, and the supernatants were collected. Total ATP levels were assessed using an ATP bioluminescence assay kit (Sigma, St. Louis, MO), following the instructions provided by the manufacturer. ATP levels were normalized against tissue weights or protein contents in the samples.

Measurement of respiration in isolated mitochondria

Mitochondria were freshly isolated from skeletal muscle and heart tissues. Respiration of mitochondria was measured using a Seahorse XF24 analyzer as described previously 39. Mitochondria (10 µg of protein) were plated in each well of a XF24 plate in 50 μl 1x Mitochondrial Assay Solution (MAS), pH 7.4 [Sucrose (70 mM), Mannitol (220 mM), KH2PO4 (5 mM), MgCl2 (5 mM), HEPES (2 mM), EGTA (1 mM), and fatty acids-free BSA (0.2%)]. XF24 plates were centrifuged at 4°C for 20 min at 2,000g, and then 450 μl of MAS containing 5 mM pyruvate/5 mM malate, 5 mM glutamate/5 mM malate, 40 μM palmitoyl-CoA/40 μM carnitine/5 mM malate, or 10 mM succinate was added to each well. Plates were incubated at 37°C in an incubator for 8–10 min. Oxygen consumption rates (OCRs) were measured under basal conditions in the presence of ADP (4 mM), the ATP synthase inhibitor oligomycin (1.5 μM), the mitochondrial uncoupling compound carbonylcyanide-4-trifluorometh-oxyphenylhydrazone (FCCP, 4 μM), and the electron transport chain inhibitor rotenone (1 μM).

Measurement of respiration in fresh muscle tissues

OCRs of muscle tissues were measured using a Seahorse XF24 analyzer and Seahorse XF24 Islet Capture Microplates (Agilent, 101122-100) as described previously 40. Briefly, tibialis anterior muscles were dissected into 3 x 1 mm cross-sections with a biopsy punch (3 mm diameter). The tissue sections were incubated for 1 hour in DMEM supplemented with 10 mM glucose, 2 mM L-glutamine, and 1 mM sodium pyruvate. Tissue OCRs were measured at the basal level and following the addition of oligomycin (8 μM), FCCP (4 μM), and antimycin A/rotenone (1 μM). Following OCR measurements, tissues were dried at 60 °C for 48 hours. All tissue OCR measurements were then normalized against dry tissue weights.

Mitochondrial pyruvate dehydrogenase (PDH) activity assay, pyruvate, acetyl-CoA, and

α-ketoglutarate (α-KG) measurements

Mitochondria were freshly isolated from skeletal muscle/heart tissues and washed twice with MAS. Mitochondrial pellets were lysed in [Tris-HCl (pH 7.4, 50 mM), NP-40 (1%), Na-deoxycholate (0.25%), NaCl (150 mM), EDTA (1 mM), NaF (1 mM), Na3VO4 (1 mM), PMSF (1 mM), aprotinin/leupeptin (10 µg/ml)]. Lysates were subjected to the PDH activity assay and pyruvate, acetyl-CoA, and α-KG measurements using PDH activity colorimetric assay, pyruvate assay, acetyl-CoA assay, and α-KG assay kits purchased from BioVision following the manufacturer’s instructions.

Histological and transmission electron microscopic (TEM) examination

Freshly dissected skeletal muscle and heart tissues were embedded in Optimal Cutting Temperature (OCT) compound (Tissue-Tek) and frozen at -80°C. Slides were stained for Hematoxylin and Eosin (H&E), Masson’s trichrome, and Oil Red O. For TEM analyses, mice were perfused with 3% paraformaldehyde, 1.5% glutaraldehyde, 100 mM cacodylate (pH 7.4), and 2.5% sucrose. Skeletal muscle and heart tissues were postfixed for 1 hour and then processed for TEM examination.

Tissue reactive oxygen species (ROS) determination

Freshly dissected heart tissues were embedded in Optimal Cutting Temperature (OCT) compound (Tissue-Tek) and immediately placed on crushed dry ice. Frozen sections were prepared at 8 μm thickness. After being rinsed in pure H2O for 30 seconds to wash out OCT compound, slides were stained with 5 μM dihydroethidium (DHE) at room temperature in dark for 15 min. Slides were washed with deionized H2O for 1 min three times and visualized immediately using a fluorescence microscope. DHE-derived 2-OH-E+, a specific adduct of cellular superoxide, was visualized with a red excitation filter. Equal exposure times were applied to all slides. Fluorescence intensities were measured by ImageJ software.

Immunoblotting

Skeletal muscle and heart tissues were lysed in RIPA buffer (50 mM Tris-HCl pH 7.4, 1% NP-40, 0.25% Na-deoxycholate, 150 mM NaCl, 1 mM EDTA, 1 mM NaF, 1 mM Na3VO4, 10 μg/mL leupeptin, 10 μg/mL aprotinin, and 1 mM PMSF). Lysates (50 μg) were resolved by SDS-PAGE followed by immunoblotting with the indicated antibodies following standard procedures.

Statistics

Data are presented as mean ± S.D. of all mice analyzed (i.e., biological replicates) in multiple experiments. Tissues, mitochondria, etc. that were analyzed were isolated from single individual mice (not pooled). Statistical significance was determined using unpaired two-tailed Student’s t test. * p < 0.05; ** p < 0.01; *** p < 0.001. N.S. indicates not significant.

Supporting information